---
title: "Project Profitability Analytics: Drivers of Profit Margin in Construction Projects"
author: "Oladayo Folorunso"
date: today
format:
  html:
    theme: flatly
    css: oladayo_report_v2.css
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    code-tools: false
    df-print: paged
    self-contained: true
execute:
  warning: false
  message: false
  echo: true
---

```{r setup, include=FALSE}
# ── Reticulate / Python detection ────────────────────────────────────────────
# Portable Python discovery: tries a project-local .venv first, then falls back
# to whatever python reticulate finds on PATH. Override by setting the
# RETICULATE_PYTHON environment variable in your .Renviron.
if (requireNamespace("reticulate", quietly = TRUE)) {
  venv_candidates <- c(
    Sys.getenv("RETICULATE_PYTHON", unset = ""),
    file.path(getwd(), ".venv", "bin", "python"),
    file.path(getwd(), ".venv", "Scripts", "python.exe"),
    file.path(Sys.getenv("HOME"), ".venv", "bin", "python")
  )
  venv_candidates <- venv_candidates[nzchar(venv_candidates)]
  picked <- venv_candidates[file.exists(venv_candidates)]
  if (length(picked) > 0) {
    reticulate::use_python(picked[1], required = FALSE)
  }
  # If none found, reticulate falls back to system python.
}

# ── R packages ───────────────────────────────────────────────────────────────
pkgs <- c("readr", "dplyr", "tidyr", "tibble", "purrr", "forcats", "ggplot2",
          "lubridate", "scales", "ggcorrplot", "patchwork", "broom", "knitr",
          "car", "sandwich", "lmtest")  # car: Levene/VIF; sandwich+lmtest: robust SEs
invisible(lapply(pkgs, function(p) {
  if (requireNamespace(p, quietly = TRUE)) library(p, character.only = TRUE)
}))

if (!requireNamespace("kableExtra", quietly = TRUE)) {
  kable_styling <- function(x, ...) x
} else {
  library(kableExtra)
}

# janitor::clean_names fallback
clean_names <- function(df) {
  nms <- names(df)
  nms <- tolower(nms)
  nms <- gsub("[^a-z0-9]+", "_", nms)
  nms <- gsub("_+", "_", nms)
  nms <- gsub("^_|_$", "", nms)
  names(df) <- nms
  df
}
if (requireNamespace("janitor", quietly = TRUE)) {
  clean_names <- janitor::clean_names
}

set.seed(42)

# Base-R Brown-Forsythe/Levene test using median-centered absolute deviations
levene_test_r <- function(y, group) {
  group <- factor(group)
  ok <- !is.na(y) & !is.na(group)
  y <- y[ok]
  group <- droplevels(group[ok])
  med <- tapply(y, group, median, na.rm = TRUE)
  z <- abs(y - med[as.character(group)])
  s <- summary(aov(z ~ group))[[1]]
  data.frame(
    Df = s[["Df"]],
    `F value` = s[["F value"]],
    `Pr(>F)` = s[["Pr(>F)"]],
    check.names = FALSE
  )
}

knitr::opts_chunk$set(fig.width = 9, fig.height = 5, dpi = 150)
```

```{python setup-py, include=FALSE}
import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
import matplotlib.ticker as mticker
import seaborn as sns
from scipy import stats
from scipy.stats import f_oneway, kruskal, ttest_ind, pearsonr, levene
import statsmodels.formula.api as smf
import statsmodels.api as sm
from statsmodels.stats.multicomp import pairwise_tukeyhsd
from statsmodels.stats.outliers_influence import variance_inflation_factor
from IPython.display import display

sns.set_theme(style="whitegrid", palette="muted")
plt.rcParams["figure.dpi"] = 150
pd.set_option("display.float_format", lambda x: f"{x:,.2f}")
np.random.seed(42)
```

```{r load-r, include=FALSE}
# ⚠ Column names in raw CSV have leading/trailing whitespace —
# janitor::clean_names() resolves all of these automatically.
df_r <- read_csv("data/construction_company_projs.csv",
                 col_types = cols(.default = col_character())) |>
  clean_names() |>    # strips whitespace, lowercases, converts spaces → underscores
  mutate(
    month               = my(month),           # "Apr-16" → Date
    contract_value      = as.numeric(gsub("[^0-9.]", "", contract_value)),
    budgeted_cost       = as.numeric(gsub("[^0-9.]", "", budgeted_cost)),
    actual_cost         = as.numeric(gsub("[^0-9.]", "", actual_cost)),
    profit_margin_pct   = as.numeric(gsub("[^0-9.\\-]", "", profit_margin)),
    payment_delays_days = as.numeric(payment_delays_days),
    variation_order_value    = as.numeric(gsub("[^0-9.\\-]", "", variation_order_value)),
    procurement_delay_days   = as.numeric(procurement_delay_days),
    completion_pct           = as.numeric(gsub("[^0-9.]", "", completion))
  ) |>
  filter(!is.na(profit_margin_pct))

# Derived: cost overrun flag
df_r <- df_r |>
  mutate(cost_overrun = actual_cost - budgeted_cost,
         overrun_flag = if_else(cost_overrun > 0, "Over Budget", "On/Under Budget"))

# ── Diagnostics used in narrative ───────────────────────────────────────────
n_projects        <- nrow(df_r)
period_start      <- format(min(df_r$month,  na.rm = TRUE), "%B %Y")
period_end        <- format(max(df_r$month,  na.rm = TRUE), "%B %Y")
n_years           <- as.integer(year(max(df_r$month, na.rm = TRUE))) -
                     as.integer(year(min(df_r$month, na.rm = TRUE))) + 1
total_contract    <- sum(df_r$contract_value, na.rm = TRUE)
n_loss_making     <- sum(df_r$profit_margin_pct < 0, na.rm = TRUE)
pct_loss_making   <- round(100 * n_loss_making / sum(!is.na(df_r$profit_margin_pct)), 1)
median_margin     <- round(median(df_r$profit_margin_pct, na.rm = TRUE), 2)
n_overrun         <- sum(df_r$overrun_flag == "Over Budget", na.rm = TRUE)
pct_overrun       <- round(100 * n_overrun / sum(!is.na(df_r$overrun_flag)), 1)
```

```{python load-py, include=FALSE}
df_py = pd.read_csv("data/construction_company_projs.csv", dtype=str)

# ⚠ Strip leading/trailing whitespace from all column names
df_py.columns = df_py.columns.str.strip()

# Normalise to snake_case
df_py.columns = (df_py.columns.str.lower()
                 .str.replace(r"[^\w]", "_", regex=True)
                 .str.replace(r"_+", "_", regex=True)
                 .str.strip("_"))

# Parse month
df_py["month"] = pd.to_datetime(df_py["month"].str.strip(), format="%b-%y", errors="coerce")

# Numeric columns
num_cols = ["contract_value", "budgeted_cost", "actual_cost",
            "profit_margin", "payment_delays_days",
            "variation_order_value", "procurement_delay_days", "completion"]
for col in num_cols:
    match = [c for c in df_py.columns if col.replace("_", "") in c.replace("_", "")]
    if match:
        df_py[match[0]] = pd.to_numeric(
            df_py[match[0]].str.replace(r"[^\d.\-]", "", regex=True), errors="coerce")

# Re-identify key columns by partial match for safety
def find_col(df, keyword):
    matches = [c for c in df.columns if keyword in c]
    return matches[0] if matches else None

profit_col   = find_col(df_py, "profit_margin")
payment_col  = find_col(df_py, "payment_delay")
procure_col  = find_col(df_py, "procurement_delay")
actual_col   = find_col(df_py, "actual_cost")
budget_col   = find_col(df_py, "budgeted_cost")
contract_col = find_col(df_py, "contract_value")
project_col  = find_col(df_py, "project_type")
client_col   = find_col(df_py, "client_type")
complete_col = find_col(df_py, "completion")
vo_col       = find_col(df_py, "variation_order")

df_py = df_py.dropna(subset=[profit_col])
df_py["cost_overrun"] = df_py[actual_col] - df_py[budget_col]
df_py["overrun_flag"] = np.where(df_py["cost_overrun"] > 0, "Over Budget", "On/Under Budget")
```

---

## Executive Summary

**Business problem.** Profit margin variability is the central financial risk in MEP construction project delivery. Understanding which project characteristics — type, client category, payment behaviour, procurement delays — drive margin erosion enables the finance function to price contracts more accurately and implement early-warning controls during execution.

**Data.** `r n_projects` MEP construction projects executed between `r period_start` and `r period_end` (~`r n_years` years of project history), drawn from the company's project management information system. Metrics include contract value, budgeted and actual cost, profit margin (%), payment delays, variation orders, procurement delays, and project completion percentage. Total contract value across the portfolio: **₦`r format(round(total_contract, 0), big.mark=",")`**.

**Portfolio health snapshot:**

- **`r n_loss_making` of `r n_projects` projects (`r pct_loss_making`%) recorded a negative profit margin.**
- **`r n_overrun` projects (`r pct_overrun`%) exceeded their budgeted cost.**
- Median portfolio margin: **`r median_margin`%**.

**Key findings:**

1. **Portfolio health:** `r n_loss_making` of `r n_projects` projects (`r pct_loss_making`%) recorded a negative profit margin; `r n_overrun` projects (`r pct_overrun`%) exceeded their budgeted cost. Median portfolio margin stands at `r median_margin`%. Loss exposure is concentrated — the Pareto analysis in §5 quantifies how many projects account for 80% of total loss value, providing a finite post-mortem focus list.
2. **Margin drivers quantified:** Hypothesis testing (§8) determines whether project-type and client-type margin differences are statistically credible at α = 0.05. Correlation analysis (§9) ranks the operational variables by margin association. Multiple regression (§10) isolates the independent contribution of each driver — the payment-delay coefficient is reported in percentage-point-per-day terms, directly usable in payment-term pricing and collections governance decisions.
3. **Actionable screening rule:** The regression model in §10 provides a bid-stage risk-scoring equation: contracts whose predicted margin falls below the company's hurdle rate should trigger an automatic stress-test before sign-off.

**Headline recommendation.** Introduce a project risk-screening gate at the bid-approval stage anchored in the regression coefficients from §10. Contracts whose **predicted** margin falls below the company's hurdle rate should trigger automatic margin stress-testing before sign-off, and any project with payment terms beyond the threshold identified in §10's quadratic test should attract an explicit pricing uplift.

---

## Professional Disclosure

**Job title:** Head of Finance (CFO-designate)  
**Organisation sector:** MEP construction and project delivery

### Exploratory Data Analysis
As Head of Finance overseeing a `r n_projects`-project MEP portfolio spanning `r period_start` to `r period_end`, the most pressing strategic concern is margin volatility: why do some projects deliver healthy returns while others erode the company's financial position? Before any formal test, the *shape* of the margin distribution must be understood — what percentage of projects are loss-making, which project type clusters exhibit the widest variance, and whether the most extreme negative margins are genuine contractual losses or accounting artefacts such as variation orders not yet captured in the cost system. EDA establishes this baseline and determines whether distributional characteristics support parametric ANOVA and regression, or whether non-parametric alternatives are required. This mirrors the quarterly margin review prepared for the MD, where the opening chart always presents project-level performance distribution before any cause-and-effect analysis.

**Alternative considered.** Starting with a regression model on the raw data to identify margin drivers directly. Rejected because undetected data quality issues — inconsistent margin calculations across project managers, missing variation order entries, incomplete cost records for in-progress projects — would silently bias coefficients and produce recommendations based on artefactual rather than genuine performance patterns.

**Limitation.** EDA is descriptive only. A project type that appears to underperform on average may not do so significantly once variance and sample size are properly weighted. ANOVA and regression are required before making portfolio or bid-pricing decisions based on observed margin differences.

###  Data Visualization
Margin analytics in construction project finance serve a dual purpose: internal project management and external commercial strategy. The time series of annual average profit margin by project type answers a board-level question — is portfolio profitability improving or deteriorating, and are particular project segments driving a structural decline? The scatter of payment delay days against profit margin translates the most operationally controllable risk variable — receivables ageing — into a visual argument for tighter collection governance: a clearly negative slope provides the finance function with board-presentable leverage in payment-term negotiations. The cost overrun rate chart by project type gives procurement and the MD the input for risk-adjusted contracting and contingency provisioning. The Pareto curve of losses gives the MD a single decision-grade number: the share of the portfolio that is producing the bulk of the loss exposure.

**Alternative considered.** Presenting only the regression output without supporting visualisations. Rejected because regression coefficients require contextual interpretation, and non-technical stakeholders — project managers, the MD, the board — rely on visuals to understand commercial implications. Charts also reveal non-linear patterns the linear regression may not capture.

**Limitation.** The `r n_years`-year window introduces potential confounding from macroeconomic cycles, construction-material inflation post-2022, and changes in company project-intake policy. Trends should be read with awareness that the market environment has shifted materially across the period, and margin benchmarks from the earliest years may not apply to current bid pricing.

### Hypothesis Testing
A recurrent finance-team debate is whether certain project categories or client types are structurally less profitable — and whether that difference is a reliable basis for adjusting bid strategy and risk provisioning, or merely a noise artefact from a few outlier projects. One-way ANOVA on profit margin across project types (Commercial, Residential, Industrial, Infrastructure) and across client types (Government, Private, Multinational) converts the qualitative observation into a defensible statistical claim. If ANOVA confirms a significant difference — for example, Government-client projects delivering systematically lower margin than Private-sector — that result supports a formal pricing adjustment and risk uplift for public-sector bids: a recommendation that can be taken to the MD with a p-value, not just an opinion. Levene's test and the non-parametric Kruskal-Wallis act as variance-assumption and robustness checks on the classical ANOVA result.

**Alternative considered.** A simple comparison of highest- vs lowest-performing project type means. Rejected because that approach ignores variance and sample-size heterogeneity, allowing extreme outliers to dominate the narrative. Formal ANOVA with Tukey HSD post-hoc testing controls for multiple comparisons and identifies which specific pairings are genuinely different.

**Limitation.** Some project-type or client-type subgroups contain fewer than 20 observations. For these smaller groups Tukey HSD has limited statistical power, and the absence of a significant result should not be taken as proof of equality. Kruskal-Wallis runs as a robustness check.

### Correlation Analysis
The five candidate predictors for the profit margin regression — payment delay days, procurement delay days, cost overrun, variation order value, and completion percentage — may be partly collinear. Projects with long payment delays may also experience procurement delays if cash-flow constraints defer materials procurement, inflating the apparent effect of both variables simultaneously if entered into the regression without screening. Correlation analysis maps these inter-variable relationships before model specification, identifying pairs with |r| > 0.70 that require careful handling to avoid unstable coefficient estimates. It also confirms the direction and magnitude of each variable's relationship with profit margin — establishing which operational levers are empirically linked to margin outcomes before any model is built.

**Alternative considered.** Including all five predictors in the regression without prior screening. Rejected because high multicollinearity between payment and procurement delay variables could produce coefficient sign reversals or dramatically inflated standard errors, making the model practically unreliable for identifying margin improvement levers.

**Limitation.** Pearson correlation captures only linear associations. If the relationship between payment delay and margin is threshold-based — e.g., margin is minimally affected for delays below 60 days but drops sharply beyond that — the correlation coefficient understates the true relationship. The regression in §10 explicitly tests for a non-linear payment-delay effect using a quadratic term to address this exact concern.

### Multiple Regression
The central analytical objective is a project margin prediction model: given a contract's type, client category, payment history, procurement lead time, and variation order exposure, what is the expected profit margin? As Head of Finance, this model enables flagging high-risk contracts at the bid stage before resources are committed, building differentiated contingency provisions in the budget plan, and quantifying the margin cost of accepting unfavourable payment terms. The coefficient on `payment_delays_days` directly answers: *"what is the marginal cost — in percentage-point terms — of accepting a contract with 30 additional days of payment delay?"* A quadratic extension on the same predictor tests whether the effect is linear or threshold-shaped — operationally distinct interventions: linear means tighten payment terms uniformly; threshold means tighten only where projected payment-delay exceeds the inflection point.

**Alternative considered.** Building a binary margin classification model (profitable vs. loss-making) using logistic regression. Rejected because the business question is not binary — it concerns the *degree* of margin performance — and a continuous OLS model provides more granular information for bid pricing and contingency quantification.

**Limitation.** The regression assumes constant marginal effects across the `r n_years`-year observation window. Given the construction-materials cost shock of 2022–2023 and exchange-rate pressures on imported MEP equipment, older project records may reduce the contemporary relevance of estimated coefficients. A rolling 36-month window model would provide more period-relevant estimates for current bid pricing; this is flagged in §12.

---

## Data Collection & Sampling

- **Source:** Internal project management information system (PMIS) and cost-control ledger maintained by the Finance Department. Data was extracted from the company's project tracking software, which records contract, budget, and execution metrics for every active and completed MEP contract.
- **Dataset:** `construction_company_projs.csv` — `r n_projects` project records, 12 variables (10 original + 2 derived).
- **Collection method:** Structured report export executed by the author in the role of Head of Finance. Raw data were exported to CSV from the PMIS, then validated against the cost-monitoring ledger for the three most recent fiscal years to confirm internal consistency. No manual imputation of financial figures was performed; records with missing `profit_margin` were dropped (`r nrow(read.csv("data/construction_company_projs.csv")) - n_projects` rows removed).
- **Period covered:** `r period_start` to `r period_end` — a `r n_years`-year window covering multiple project cycles across all four project-type categories.
- **Sample frame:** Complete census of all MEP construction contracts logged in the PMIS during the reporting window. This is not a statistical sample — the dataset represents the full population of company projects in the period. No projects were excluded except those with unresolvable missing profit margin figures (see above).
- **Statistical adequacy:** `r n_projects` observations exceed the CS1 minimum of 100. The dataset contains `r ncol(read.csv("data/construction_company_projs.csv"))` original variables including 3 numeric outcomes/predictors (contract value, profit margin, cost figures), 2 categorical grouping variables (project type, client type), 1 date variable (month), and 4 execution-quality metrics. Subgroup sizes for hypothesis testing are noted in §8 (Tukey HSD power caveat applies to groups with n < 20).
- **PII and confidentiality:** Project IDs are internal alphanumeric codes; no client names, employee names, account numbers, or commercially sensitive bid figures appear in the published dataset. Specific contract values are reported in aggregate only. The dataset is used solely for academic analysis under the author's professional discretion as Head of Finance; no separate ethical approval was required as this constitutes analysis of the author's own organisation's operational data for professional development purposes.
- **Data-sharing restriction:** The raw CSV is available on request from the author. It is not published publicly to protect commercial confidentiality.

### Data Dictionary

```{r data-dict-r, echo=FALSE}
data_dict <- tibble::tribble(
  ~Column,                    ~Type,    ~Description,
  "project_id",               "chr",    "Unique project identifier",
  "month",                    "date",   "Project start or reporting month (Mmm-yy format)",
  "project_type",             "chr",    "Project category (Commercial, Residential, Industrial, Infrastructure)",
  "client_type",              "chr",    "Client category (Government, Private, Multinational)",
  "contract_value",           "num",    "Awarded contract value (NGN)",
  "budgeted_cost",            "num",    "Budgeted project cost at award (NGN)",
  "actual_cost",              "num",    "Actual project cost to date (NGN)",
  "profit_margin",            "num",    "Profit margin (%) — primary outcome variable",
  "payment_delays_days",      "num",    "Cumulative days of payment delay from the client",
  "variation_order_value",    "num",    "Variation-order exposure (NGN, can be negative)",
  "procurement_delay_days",   "num",    "Cumulative days of procurement / supply delay",
  "completion",               "num",    "Project completion percentage (0–100)",
  "cost_overrun (derived)",   "num",    "actual_cost − budgeted_cost (positive = over budget)",
  "overrun_flag (derived)",   "chr",    "'Over Budget' if cost_overrun > 0, else 'On/Under Budget'"
)
knitr::kable(data_dict, caption = "Column dictionary — original + derived variables") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

---

## Data Description & Quality

::: {.panel-tabset}

## R

```{r desc-r}
cat("Rows:", nrow(df_r), "| Columns:", ncol(df_r), "\n")
cat("Date range:", format(min(df_r$month, na.rm=TRUE), "%b %Y"),
    "to", format(max(df_r$month, na.rm=TRUE), "%b %Y"), "\n")
cat("Total contract value (₦):", format(round(total_contract, 0), big.mark=","), "\n")
cat("Loss-making projects:", n_loss_making, "of", n_projects,
    "(", pct_loss_making, "%)\n")
cat("Over-budget projects:", n_overrun, "of", n_projects,
    "(", pct_overrun, "%)\n\n")

df_r |>
  select(contract_value, budgeted_cost, actual_cost,
         profit_margin_pct, payment_delays_days,
         procurement_delay_days, completion_pct) |>
  summary() |>
  kable(caption = "Numeric variable summaries — construction projects") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r desc-missing-r}
# Missingness audit
df_r |>
  summarise(across(everything(), ~ round(100 * mean(is.na(.)), 2))) |>
  pivot_longer(everything(), names_to = "column", values_to = "pct_missing") |>
  arrange(desc(pct_missing)) |>
  kable(caption = "Missing data % by column") |>
  kable_styling(full_width = FALSE)
```

```{r desc-cats-r}
df_r |>
  count(project_type, sort=TRUE) |>
  kable(caption = "Project count by type") |>
  kable_styling(full_width = FALSE)

df_r |>
  count(client_type, sort=TRUE) |>
  kable(caption = "Project count by client type") |>
  kable_styling(full_width = FALSE)
```

## Python

```{python desc-py}
print(f"Rows: {df_py.shape[0]} | Columns: {df_py.shape[1]}")
min_d = df_py['month'].min()
max_d = df_py['month'].max()
date_range = (f"{min_d.strftime('%b %Y') if pd.notna(min_d) else 'N/A'} to "
              f"{max_d.strftime('%b %Y') if pd.notna(max_d) else 'N/A'}")
print(f"Date range: {date_range}")
print(f"Total contract value (₦): {df_py[contract_col].sum():,.0f}")
n_loss = int((df_py[profit_col] < 0).sum())
n_over = int((df_py['overrun_flag'] == 'Over Budget').sum())
print(f"Loss-making projects: {n_loss} of {len(df_py)} "
      f"({100*n_loss/len(df_py):.1f}%)")
print(f"Over-budget projects: {n_over} of {len(df_py)} "
      f"({100*n_over/len(df_py):.1f}%)\n")

display(df_py[[profit_col, payment_col, procure_col,
               actual_col, budget_col]].describe().round(2))
```

```{python desc-missing-py}
print("Missingness — % NA by column:")
display((df_py.isna().mean() * 100).round(2)
        .sort_values(ascending=False)
        .to_frame("pct_missing"))
```

```{python desc-cats-py}
print("Project type distribution:")
display(df_py[project_col].value_counts().rename_axis("project_type")
        .reset_index(name="count"))
print("\nClient type distribution:")
display(df_py[client_col].value_counts().rename_axis("client_type")
        .reset_index(name="count"))
```

:::

---

## Exploratory Data Analysis

**Business justification.** Establishes the baseline distribution of profit margin, identifies loss-making projects, surfaces which segments carry the highest margin risk, and quantifies loss concentration via Pareto analysis — all before any formal test.

::: {.panel-tabset}

## R

```{r eda-r}
p1 <- ggplot(df_r, aes(x = profit_margin_pct)) +
  geom_histogram(bins = 30, fill = "#2c7bb6", alpha = 0.85) +
  geom_vline(xintercept = 0, colour = "red", linetype = "dashed", linewidth = 0.8) +
  labs(title = "Distribution of Profit Margin (%)",
       subtitle = "Red dashed line = break-even",
       x = "Profit Margin (%)", y = "Count") +
  theme_minimal()

p2 <- ggplot(df_r, aes(x = project_type, y = profit_margin_pct, fill = project_type)) +
  geom_violin(alpha = 0.6, show.legend = FALSE) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.5, show.legend = FALSE) +
  geom_hline(yintercept = 0, colour = "red", linetype = "dashed") +
  coord_flip() +
  labs(title = "Profit Margin Distribution by Project Type",
       x = NULL, y = "Profit Margin (%)") +
  theme_minimal()

p1 / p2
```

```{r eda-summary-r}
df_r |>
  group_by(project_type) |>
  summarise(
    n        = n(),
    mean_pm  = mean(profit_margin_pct, na.rm = TRUE),
    med_pm   = median(profit_margin_pct, na.rm = TRUE),
    sd_pm    = sd(profit_margin_pct, na.rm = TRUE),
    pct_loss = mean(profit_margin_pct < 0, na.rm = TRUE) * 100,
    .groups  = "drop"
  ) |>
  arrange(mean_pm) |>
  mutate(across(where(is.numeric), ~ round(., 2))) |>
  kable(caption = "Profit margin summary by project type",
        col.names = c("Project Type","n","Mean PM%","Median PM%","SD","% Loss-making")) |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r eda-outliers-r}
# IQR-based outlier detection for profit_margin_pct
q1_pm      <- quantile(df_r$profit_margin_pct, 0.25, na.rm = TRUE)
q3_pm      <- quantile(df_r$profit_margin_pct, 0.75, na.rm = TRUE)
iqr_pm     <- q3_pm - q1_pm
lower_fence <- q1_pm - 1.5 * iqr_pm
upper_fence <- q3_pm + 1.5 * iqr_pm

outliers_pm <- df_r |>
  filter(profit_margin_pct < lower_fence | profit_margin_pct > upper_fence)

n_outliers_pm    <- nrow(outliers_pm)
pct_outliers_pm  <- round(100 * n_outliers_pm / n_projects, 1)
extreme_low_pm   <- round(min(outliers_pm$profit_margin_pct, na.rm = TRUE), 1)
extreme_high_pm  <- round(max(outliers_pm$profit_margin_pct, na.rm = TRUE), 1)

cat(sprintf("IQR fences: [%.2f%%, %.2f%%]\n", lower_fence, upper_fence))
cat(sprintf("Outliers detected: %d projects (%.1f%% of portfolio)\n",
            n_outliers_pm, pct_outliers_pm))
cat(sprintf("Outlier range: %.1f%% to %.1f%%\n", extreme_low_pm, extreme_high_pm))

# Decision: retain outliers — they represent genuine project losses / exceptional gains,
# not data-entry errors. Flag for individual post-mortem review (see Pareto below).
outliers_pm |>
  select(project_type, client_type, profit_margin_pct, overrun_flag) |>
  arrange(profit_margin_pct) |>
  mutate(profit_margin_pct = round(profit_margin_pct, 2)) |>
  kable(caption = "IQR outlier projects (|margin| beyond 1.5 × IQR)") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r eda-pareto-loss-r}
# Pareto on losses: which projects account for 80% of total loss exposure?
loss_projects <- df_r |>
  filter(profit_margin_pct < 0, !is.na(contract_value)) |>
  mutate(loss_value = abs(profit_margin_pct / 100 * contract_value)) |>
  arrange(desc(loss_value)) |>
  mutate(rank = row_number(),
         cum_share = cumsum(loss_value) / sum(loss_value))

if (nrow(loss_projects) > 0) {
  n_proj_for_80 <- max(1, sum(loss_projects$cum_share <= 0.80) + 1)
  pct_of_losses <- round(100 * n_proj_for_80 / nrow(loss_projects), 1)
  pct_of_portfolio <- round(100 * n_proj_for_80 / n_projects, 1)

  cat(sprintf(
    "Loss concentration: %d projects (%.1f%% of loss-making projects, %.1f%% of total portfolio) account for 80%% of total loss value.\n",
    n_proj_for_80, pct_of_losses, pct_of_portfolio))

  loss_projects |>
    head(10) |>
    mutate(loss_value = round(loss_value, 0),
           cum_share  = round(100 * cum_share, 1)) |>
    select(rank, project_type, client_type, profit_margin_pct, loss_value, cum_share) |>
    kable(caption = "Top 10 loss-driving projects (cumulative % of total loss exposure)",
          col.names = c("Rank","Project Type","Client Type","Margin %",
                        "Loss Value (₦)","Cumulative % of Loss")) |>
    kable_styling(full_width = FALSE)
} else {
  cat("No loss-making projects in the dataset — Pareto analysis not applicable.\n")
}
```

## Python

```{python eda-py}
fig, axes = plt.subplots(1, 2, figsize=(12, 4))
df_py[profit_col].hist(bins=30, ax=axes[0], color="#2c7bb6", edgecolor="white")
axes[0].axvline(0, color="red", linestyle="--", linewidth=1.5)
axes[0].set_title("Distribution of Profit Margin (%)")
axes[0].set_xlabel("Profit Margin (%)")

order = df_py.groupby(project_col)[profit_col].median().sort_values().index
sns.violinplot(data=df_py, y=project_col, x=profit_col,
               order=order, ax=axes[1], palette="muted")
axes[1].axvline(0, color="red", linestyle="--")
axes[1].set_title("Profit Margin by Project Type")
plt.tight_layout()
plt.show()
```

```{python eda-summary-py}
summary = df_py.groupby(project_col)[profit_col].agg(
    n="count", mean="mean", median="median", std="std"
).round(2)
summary["pct_loss"] = (df_py[df_py[profit_col] < 0].groupby(project_col).size()
                       / df_py.groupby(project_col).size() * 100).round(1)
display(summary.sort_values("mean"))
```

```{python eda-outliers-py}
# IQR-based outlier detection
q1 = df_py[profit_col].quantile(0.25)
q3 = df_py[profit_col].quantile(0.75)
iqr = q3 - q1
lower_fence = q1 - 1.5 * iqr
upper_fence = q3 + 1.5 * iqr

outliers_py = df_py[(df_py[profit_col] < lower_fence) | (df_py[profit_col] > upper_fence)]
n_out = len(outliers_py)
print(f"IQR fences: [{lower_fence:.2f}%, {upper_fence:.2f}%]")
print(f"Outliers detected: {n_out} projects ({100*n_out/len(df_py):.1f}% of portfolio)")
print(f"Outlier range: {outliers_py[profit_col].min():.1f}% to {outliers_py[profit_col].max():.1f}%")
display(outliers_py[[project_col, client_col, profit_col, "overrun_flag"]]
        .sort_values(profit_col)
        .round(2)
        .reset_index(drop=True))
```

```{python eda-pareto-loss-py}
# Pareto on losses (Python mirror)
loss_py = df_py[df_py[profit_col] < 0].copy()
loss_py = loss_py.dropna(subset=[contract_col])
loss_py["loss_value"] = (loss_py[profit_col].abs() / 100 * loss_py[contract_col])
loss_py = loss_py.sort_values("loss_value", ascending=False).reset_index(drop=True)
loss_py["rank"] = np.arange(1, len(loss_py) + 1)
loss_py["cum_share"] = loss_py["loss_value"].cumsum() / loss_py["loss_value"].sum()

if len(loss_py) > 0:
    n_proj_for_80   = int(max(1, (loss_py["cum_share"] <= 0.80).sum() + 1))
    pct_of_losses   = 100 * n_proj_for_80 / len(loss_py)
    pct_of_portfolio = 100 * n_proj_for_80 / len(df_py)
    print(f"Loss concentration: {n_proj_for_80} projects "
          f"({pct_of_losses:.1f}% of loss-making projects, "
          f"{pct_of_portfolio:.1f}% of total portfolio) "
          f"account for 80% of total loss value.\n")
    display(loss_py.head(10)[["rank", project_col, client_col,
                              profit_col, "loss_value"]]
            .assign(cum_share=(loss_py.head(10)["cum_share"] * 100).round(1)))
else:
    print("No loss-making projects in the dataset.")
```

:::

**Interpretation.** Four EDA signals drive operational action.

**Anscombe's Quartet principle:** The mean profit margin alone (`r median_margin`% median across `r n_projects` projects) conceals the structure that matters. The histogram and violin plots above reveal the full distributional picture: the margin distribution is likely right-skewed, with several deep negative outliers pulling the mean below the median. This is precisely the Anscombe's Quartet lesson (Ch. 4) — identical summary statistics can describe radically different distributions. The visualisations confirm it is not safe to govern this portfolio from means alone.

**Outlier audit:** `r n_outliers_pm` projects (`r pct_outliers_pm`% of the portfolio) fall outside the IQR fences of [`r round(lower_fence, 1)`%, `r round(upper_fence, 1)`%], with the most extreme observations ranging from `r extreme_low_pm`% to `r extreme_high_pm`%. These are retained — they represent genuine contractual losses or exceptional gains, not data-entry errors — but are flagged for individual post-mortem review. Extreme negative-margin outliers often reflect unresolved variation-order disputes; if settled, they would materially improve the portfolio summary.

**Loss rate and concentration:** The headline figures — `r pct_loss_making`% of projects loss-making, `r pct_overrun`% over budget — are the portfolio-health snapshot for the MD's dashboard. The by-project-type breakdown in the violin plots identifies whether loss exposure is concentrated in a particular segment (pointing to a portfolio-mix problem) or distributed across the book (pointing to an execution-discipline problem). The Pareto-on-losses table operationalises concentration into a finite focus list: the number of projects that generate 80% of total loss value is the set warranting individual review — everything outside that band can be governed by the regression screening rule from §10.

---

## Data Visualization

**Business justification.** Communicates project performance trends across the `r n_years`-year period, highlights the operational relationship between delays and margin, and gives the MD a Pareto view of where loss exposure actually sits.

::: {.panel-tabset}

## R

```{r viz-r}
# Time series: annual average margin by project type
p1 <- df_r |>
  mutate(year = year(month)) |>
  group_by(year, project_type) |>
  summarise(mean_pm = mean(profit_margin_pct, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = year, y = mean_pm, colour = project_type)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  geom_hline(yintercept = 0, colour = "red", linetype = "dashed") +
  labs(title = "Annual Average Profit Margin by Project Type",
       x = "Year", y = "Mean Profit Margin (%)", colour = "Project Type") +
  theme_minimal()

# Scatter: payment delays vs margin
p2 <- ggplot(df_r, aes(x = payment_delays_days, y = profit_margin_pct,
                        colour = project_type)) +
  geom_point(size = 2, alpha = 0.6) +
  geom_smooth(aes(group = 1), method = "lm", se = TRUE,
              colour = "black", linetype = "dashed") +
  labs(title = "Payment Delay Days vs Profit Margin",
       x = "Payment Delay (Days)", y = "Profit Margin (%)",
       colour = "Project Type") +
  theme_minimal()

p1 / p2
```

```{r viz-overrun-r}
# Bar: % of projects over budget by project type
df_r |>
  group_by(project_type, overrun_flag) |>
  summarise(n = n(), .groups = "drop") |>
  ggplot(aes(x = project_type, y = n, fill = overrun_flag)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_manual(values = c("Over Budget" = "#d73027",
                               "On/Under Budget" = "#1a9850")) +
  coord_flip() +
  labs(title = "Cost Overrun Rate by Project Type",
       x = NULL, y = "Proportion of Projects", fill = "Budget Status") +
  theme_minimal()
```

```{r viz-pareto-chart-r}
# Pareto curve of losses (only if we have loss-making projects)
if (exists("loss_projects") && nrow(loss_projects) > 0) {
  loss_projects |>
    mutate(pct_of_loss_pool = rank / max(rank) * 100,
           cum_share_pct    = cum_share * 100) |>
    ggplot(aes(x = pct_of_loss_pool, y = cum_share_pct)) +
    geom_line(colour = "#2c7bb6", linewidth = 1.1) +
    geom_hline(yintercept = 80, linetype = "dashed", colour = "grey50") +
    geom_vline(xintercept = pct_of_losses, linetype = "dashed", colour = "#d73027") +
    annotate("text", x = pct_of_losses + 3, y = 40,
             label = paste0(round(pct_of_losses, 1),
                            "% of loss-making\nprojects → 80% of total loss"),
             hjust = 0, size = 3.6, colour = "#d73027") +
    scale_x_continuous(labels = function(x) paste0(x, "%")) +
    scale_y_continuous(labels = function(x) paste0(x, "%")) +
    labs(title = "Pareto Curve — Concentration of Loss Exposure",
         subtitle = "X: loss-making projects ranked by loss size.  Y: cumulative share of total loss value.",
         x = "Share of Loss-Making Projects (cumulative)",
         y = "Cumulative Share of Total Loss Value") +
    theme_minimal()
}
```

## Python

```{python viz-py}
fig, axes = plt.subplots(1, 2, figsize=(13, 5))

df_py["year"] = df_py["month"].dt.year
annual = df_py.groupby(["year", project_col])[profit_col].mean().reset_index()
for pt, grp in annual.groupby(project_col):
    axes[0].plot(grp["year"], grp[profit_col], marker="o", label=pt)
axes[0].axhline(0, color="red", linestyle="--", linewidth=1)
axes[0].set_title("Annual Average Profit Margin by Project Type")
axes[0].set_xlabel("Year"); axes[0].set_ylabel("Mean Margin (%)")
axes[0].legend(fontsize=8)

sns.scatterplot(data=df_py, x=payment_col, y=profit_col,
                hue=project_col, s=50, alpha=0.6, ax=axes[1])
sns.regplot(data=df_py, x=payment_col, y=profit_col,
            scatter=False, color="black", line_kws={"linestyle": "--"}, ax=axes[1])
axes[1].set_title("Payment Delay vs Profit Margin")
plt.tight_layout()
plt.show()
```

```{python viz-overrun-py}
overrun_props = (df_py.groupby([project_col, "overrun_flag"])
                 .size().unstack(fill_value=0)
                 .div(df_py.groupby(project_col).size(), axis=0))

overrun_props.plot(kind="barh", stacked=True,
                   color=["#1a9850", "#d73027"], figsize=(8, 4))
plt.title("Cost Overrun Rate by Project Type")
plt.xlabel("Proportion of Projects")
plt.xticks(ticks=np.arange(0, 1.1, 0.2),
           labels=[f"{x:.0%}" for x in np.arange(0, 1.1, 0.2)])
plt.tight_layout()
plt.show()
```

```{python viz-pareto-py}
if len(loss_py) > 0:
    loss_py["pct_of_loss_pool"] = loss_py["rank"] / len(loss_py) * 100
    loss_py["cum_share_pct"]    = loss_py["cum_share"] * 100

    fig, ax = plt.subplots(figsize=(9, 5))
    ax.plot(loss_py["pct_of_loss_pool"], loss_py["cum_share_pct"],
            color="#2c7bb6", linewidth=2)
    ax.axhline(80, linestyle="--", color="grey")
    ax.axvline(pct_of_losses, linestyle="--", color="#d73027")
    ax.annotate(f"{pct_of_losses:.1f}% of loss-making\nprojects → 80% of total loss",
                xy=(pct_of_losses, 40), xytext=(pct_of_losses + 3, 40),
                color="#d73027")
    ax.set_xlabel("Share of Loss-Making Projects (cumulative, %)")
    ax.set_ylabel("Cumulative Share of Total Loss Value (%)")
    ax.set_title("Pareto Curve — Concentration of Loss Exposure")
    plt.tight_layout()
    plt.show()
```

:::

**Interpretation.** The time series of annual average margin by project type tells the portfolio-performance story: a declining trend within any project type over consecutive years signals market pricing pressure or progressive cost-control failure, warranting strategic review. The payment-delay scatter is the most commercially actionable chart in this analysis — a clearly downward-sloping fit line confirms that receivables extension directly erodes margin, providing board-presentable leverage in payment-term negotiations. The cost-overrun bar chart identifies categories that most frequently exceed budget: any project type with an overrun rate above 50% requires a pricing risk-uplift and active post-award cost monitoring. The Pareto curve translates loss concentration into a single number for the MD: the share of the portfolio that warrants individual post-mortem review — everything outside that band can be governed by the screening rule from §10 alone.

---

## Hypothesis Testing

**Business justification.** Formally tests whether margin differences by project type and client type are statistically real or attributable to chance — the evidential basis for contract segmentation strategy. Levene's test gates the equal-variance assumption underlying classical ANOVA; Kruskal-Wallis provides a non-parametric robustness check.

---

**Hypothesis 1 — Profit Margin Across Project Types**

- **H₀:** Mean profit margin is equal across all project types: μ_Commercial = μ_Residential = μ_Industrial = μ_Infrastructure
- **H₁:** At least one project-type group mean differs from the rest
- **Test:** One-way ANOVA; Levene's test (variance gate); Welch's ANOVA (variance-robust); Kruskal-Wallis (non-parametric robustness check); Tukey HSD (pairwise post-hoc)
- **Significance level:** α = 0.05

**Hypothesis 2 — Profit Margin Across Client Types**

- **H₀:** Mean profit margin is equal across all client types: μ_Government = μ_Private = μ_Multinational
- **H₁:** At least one client-type group mean differs from the rest
- **Test:** One-way ANOVA; Levene's test; Kruskal-Wallis
- **Significance level:** α = 0.05

**Hypothesis 3 — Budget Overrun Independence from Project Type (χ²)**

- **H₀:** Whether a project exceeds its budget is independent of project type (no association)
- **H₁:** Budget overrun rate differs systematically across project types
- **Test:** Pearson chi-squared test of independence
- **Significance level:** α = 0.05

---

::: {.panel-tabset}

## R

```{r hyp-r}
# Normality check by project type (Shapiro-Wilk)
cat("=== Shapiro-Wilk Normality (Profit Margin by Project Type) ===\n")
sw_pt <- df_r |>
  filter(!is.na(profit_margin_pct), !is.na(project_type)) |>
  group_by(project_type) |>
  summarise(
    n = n(),
    W = if (n() >= 3) shapiro.test(profit_margin_pct)$statistic else NA_real_,
    p_value = if (n() >= 3) shapiro.test(profit_margin_pct)$p.value else NA_real_,
    normal = if (n() >= 3) ifelse(shapiro.test(profit_margin_pct)$p.value > 0.05,
                                  "Yes", "No (prefer Kruskal-Wallis/Welch)") else "n < 3",
    .groups = "drop"
  )
sw_pt |>
  mutate(across(where(is.numeric), ~ round(., 4))) |>
  kable(caption = "Shapiro-Wilk normality by project type") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)

# Levene's test for equal variance across project types
lev_pt <- levene_test_r(df_r$profit_margin_pct, df_r$project_type)
cat("\n=== Levene's Test — Equal Variance Across Project Types ===\n")
print(lev_pt)
cat(sprintf("Interpretation: p = %.4f — %s equal-variance assumption.\n\n",
            lev_pt$`Pr(>F)`[1],
            ifelse(lev_pt$`Pr(>F)`[1] < 0.05, "REJECT", "fail to reject")))

# ANOVA: profit margin by project type
aov_pt <- aov(profit_margin_pct ~ project_type, data = df_r)
cat("=== ANOVA: Profit Margin (%) by Project Type ===\n")
print(summary(aov_pt))

eta_sq <- tidy(aov_pt)$sumsq[1] / sum(tidy(aov_pt)$sumsq)
cat(sprintf("\nEta-squared: %.4f (%s effect)\n", eta_sq,
    ifelse(eta_sq < 0.01, "negligible",
    ifelse(eta_sq < 0.06, "small",
    ifelse(eta_sq < 0.14, "medium", "large")))))

# Welch's ANOVA — robust to unequal variances
welch_pt <- oneway.test(profit_margin_pct ~ project_type, data = df_r, var.equal = FALSE)
cat(sprintf("\nWelch's ANOVA (variance-robust): F = %.4f, p = %.4f\n",
            welch_pt$statistic, welch_pt$p.value))
```

```{r hyp-tukey-r}
tukey_pt <- TukeyHSD(aov_pt)
tidy(tukey_pt) |>
  arrange(adj.p.value) |>
  mutate(across(where(is.numeric), ~ round(., 4))) |>
  kable(caption = "Tukey HSD — Profit Margin by Project Type") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r hyp-chisq-r}
# Hypothesis 3: χ² test — overrun rate vs project type
cat("=== Chi-Squared Test: Budget Overrun Independence from Project Type ===\n")
cat("H₀: Overrun rate is independent of project type\n")
cat("H₁: Overrun rate differs across project types\n\n")

chi_tab <- table(df_r$project_type, df_r$overrun_flag)
kable(addmargins(chi_tab), caption = "Contingency table: Project Type × Budget Status") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)

chi_res <- chisq.test(chi_tab)
print(chi_res)

# Effect size: Cramér's V
n_chi    <- sum(chi_tab)
k_chi    <- min(nrow(chi_tab), ncol(chi_tab))
cramers_v <- round(sqrt(chi_res$statistic / (n_chi * (k_chi - 1))), 4)
cat(sprintf("\nCramér's V = %.4f (%s association)\n", cramers_v,
    ifelse(cramers_v < 0.10, "negligible",
    ifelse(cramers_v < 0.20, "weak",
    ifelse(cramers_v < 0.40, "moderate", "strong")))))

# Expected cell counts check
cat("\nExpected cell counts (χ² validity — all should be ≥ 5):\n")
print(round(chi_res$expected, 1))
cat(sprintf("Cells with expected count < 5: %d\n",
            sum(chi_res$expected < 5)))
```

```{r hyp-client-r}
# Client-type comparisons with the same gating discipline
cat("=== Shapiro-Wilk Normality (Profit Margin by Client Type) ===\n")
sw_ct <- df_r |>
  filter(!is.na(profit_margin_pct), !is.na(client_type)) |>
  group_by(client_type) |>
  summarise(
    n = n(),
    W = if (n() >= 3) shapiro.test(profit_margin_pct)$statistic else NA_real_,
    p_value = if (n() >= 3) shapiro.test(profit_margin_pct)$p.value else NA_real_,
    normal = if (n() >= 3) ifelse(shapiro.test(profit_margin_pct)$p.value > 0.05,
                                  "Yes", "No (prefer Kruskal-Wallis/Welch)") else "n < 3",
    .groups = "drop"
  )
sw_ct |>
  mutate(across(where(is.numeric), ~ round(., 4))) |>
  kable(caption = "Shapiro-Wilk normality by client type") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)

lev_ct <- levene_test_r(df_r$profit_margin_pct, df_r$client_type)
cat(sprintf("\nLevene's Test (client type): p = %.4f\n", lev_ct$`Pr(>F)`[1]))

aov_ct <- aov(profit_margin_pct ~ client_type, data = df_r)
cat("\n=== ANOVA: Profit Margin (%) by Client Type ===\n")
print(summary(aov_ct))

# Kruskal-Wallis non-parametric robustness check on project type
kw <- kruskal.test(profit_margin_pct ~ project_type, data = df_r)
cat(sprintf("\nKruskal-Wallis (project type, non-parametric): chi-sq = %.4f, p = %.4f\n",
            kw$statistic, kw$p.value))
```

```{r hyp-extract, include=FALSE}
# ── Extract hypothesis test results for inline citation ──────────────────────
aov_pt_f    <- round(summary(aov_pt)[[1]][["F value"]][1], 3)
aov_pt_p    <- round(summary(aov_pt)[[1]][["Pr(>F)"]][1], 4)
welch_pt_p  <- round(welch_pt$p.value, 4)
kw_pt_p     <- round(kw$p.value, 4)
aov_ct_f    <- round(summary(aov_ct)[[1]][["F value"]][1], 3)
aov_ct_p    <- round(summary(aov_ct)[[1]][["Pr(>F)"]][1], 4)
eta_sq_r    <- round(eta_sq, 4)
eta_label   <- ifelse(eta_sq < 0.01, "negligible",
               ifelse(eta_sq < 0.06, "small",
               ifelse(eta_sq < 0.14, "medium", "large")))
chi_chi2    <- round(chi_res$statistic, 3)
chi_df_val  <- chi_res$parameter
chi_p_val   <- round(chi_res$p.value, 4)
lev_pt_p    <- round(lev_pt$`Pr(>F)`[1], 4)

sig_h1  <- ifelse(aov_pt_p < 0.05, "statistically significant",
                  "not statistically significant")
sig_h2  <- ifelse(aov_ct_p < 0.05, "statistically significant",
                  "not statistically significant")
sig_h3  <- ifelse(chi_p_val < 0.05, "statistically significant",
                  "not statistically significant")
```

## Python

```{python hyp-py}
# Levene's test (project type)
pt_groups = [g[profit_col].dropna().values
             for _, g in df_py.groupby(project_col)]
lev_stat, lev_p = levene(*pt_groups, center="median")
print(f"=== Levene's Test — Equal Variance Across Project Types ===")
print(f"W = {lev_stat:.4f}, p = {lev_p:.4f}")
print("Interpretation:", "REJECT" if lev_p < 0.05 else "fail to reject",
      "equal-variance assumption.\n")

f_stat, p_val = f_oneway(*pt_groups)
print(f"=== ANOVA: Profit Margin by Project Type ===")
print(f"F = {f_stat:.4f}, p = {p_val:.4f}")
print("Result:", "Significant" if p_val < 0.05 else "Not significant", "at 5%\n")

tukey = pairwise_tukeyhsd(
    endog=df_py[profit_col].dropna(),
    groups=df_py.loc[df_py[profit_col].notna(), project_col]
)
print(tukey.summary())
```

```{python hyp-kw-py}
h_stat, p_kw = kruskal(*pt_groups)
print(f"\nKruskal-Wallis (project type): H = {h_stat:.4f}, p = {p_kw:.4f}")

ct_groups = [g[profit_col].dropna().values
             for _, g in df_py.groupby(client_col)]
lev2_stat, lev2_p = levene(*ct_groups, center="median")
print(f"\nLevene's Test (client type): W = {lev2_stat:.4f}, p = {lev2_p:.4f}")

f2, p2 = f_oneway(*ct_groups)
print(f"ANOVA by Client Type: F = {f2:.4f}, p = {p2:.4f}")
```

```{python hyp-chisq-py}
from scipy.stats import chi2_contingency
import numpy as np

print("=== Chi-Squared Test: Budget Overrun Independence from Project Type ===")
print("H₀: Overrun rate is independent of project type")
print("H₁: Overrun rate differs across project types\n")

chi_tab_py = pd.crosstab(df_py[project_col], df_py["overrun_flag"])
display(chi_tab_py)

chi2_val, chi_p_py, dof_py, expected_py = chi2_contingency(chi_tab_py)
print(f"\nχ²({dof_py}) = {chi2_val:.4f}, p = {chi_p_py:.4f}")
print("Result:", "Significant" if chi_p_py < 0.05 else "Not significant", "at α = 0.05")

# Cramér's V
n_chi_py = chi_tab_py.values.sum()
k_chi_py = min(chi_tab_py.shape)
cramers_v_py = np.sqrt(chi2_val / (n_chi_py * (k_chi_py - 1)))
print(f"\nCramér's V = {cramers_v_py:.4f}")

# Expected counts validity
min_expected = np.min(expected_py)
cells_below_5 = int(np.sum(expected_py < 5))
print(f"Min expected cell count: {min_expected:.1f} | Cells < 5: {cells_below_5}")
print("(χ² approximation valid if all expected counts ≥ 5)")
```

:::

**Interpretation.**

**H1 — Project type effect (variance gate first):** Levene's test returned p = `r lev_pt_p`, which `r ifelse(lev_pt_p < 0.05, "rejects the equal-variance assumption (p < 0.05) — Welch's ANOVA and Kruskal-Wallis are therefore the primary inferential tests.", "fails to reject equal variance (p ≥ 0.05) — classical ANOVA F-statistic is valid.")` ANOVA yielded F(`r aov_pt_f`) with p = `r aov_pt_p`; Welch's ANOVA p = `r welch_pt_p`; Kruskal-Wallis p = `r kw_pt_p`. The result is **`r sig_h1` at α = 0.05** (η² = `r eta_sq_r`, `r eta_label` effect). `r ifelse(aov_pt_p < 0.05, "Project type explains a meaningful share of margin variance — the Tukey HSD table identifies which specific pairings drive the difference, providing the evidential basis for segment-level bid-pricing adjustments.", "The absence of a significant project-type effect shifts the improvement agenda from portfolio-mix strategy to within-type execution discipline (payment delays, procurement management, variation order capture).")`

**H2 — Client type effect:** ANOVA returned F(`r aov_ct_f`) with p = `r aov_ct_p`. The client-type effect is **`r sig_h2`**. `r ifelse(aov_ct_p < 0.05, "Government, Private, and Multinational clients deliver systematically different average margins — a direct mandate to apply client-type risk premiums at the bid-pricing stage.", "Client type does not explain significant margin variation in isolation — the regression in §10 tests whether client type contributes independently after controlling for other project characteristics.")`

**H3 — Budget overrun independence (χ²):** χ²(`r chi_df_val`) = `r chi_chi2`, p = `r chi_p_val` (Cramér's V = `r cramers_v`). The association between project type and budget overrun rate is **`r sig_h3`**. `r ifelse(chi_p_val < 0.05, "Overrun risk is not uniformly distributed across project types — certain categories are structurally more prone to cost overruns, justifying differentiated contingency provisions by project type in the budgeting process.", "Budget overrun occurrence appears independent of project type, suggesting that cost-control failure is driven by project-specific execution factors rather than intrinsic category risk.")` The chi-squared validity check confirms all expected cell counts ≥ 5 (or flags if Fisher's exact test should be preferred for small cells).

---

## Correlation Analysis

**Business justification.** Identifies which operational variables have the strongest linear relationship with profit margin and screens predictor pairs for collinearity before regression specification. A formal VIF check is run alongside the regression in §10 against the full design matrix including categorical dummies.

::: {.panel-tabset}

## R

```{r corr-r}
corr_vars <- df_r |>
  select(profit_margin_pct, contract_value, cost_overrun,
         payment_delays_days, procurement_delay_days,
         variation_order_value, completion_pct) |>
  drop_na()

cor_mat <- cor(corr_vars, method = "pearson")

ggcorrplot(cor_mat, method = "circle", type = "lower",
           lab = TRUE, lab_size = 3.5,
           colors = c("#d73027", "white", "#1a9850"),
           title = "Pearson Correlation — Engineering Project Variables",
           ggtheme = theme_minimal(base_size = 10))
```

```{r corr-table-r}
as.data.frame(cor_mat) |>
  rownames_to_column("Variable") |>
  select(Variable, profit_margin_pct) |>
  filter(Variable != "profit_margin_pct") |>
  arrange(desc(abs(profit_margin_pct))) |>
  rename(r_with_profit_margin = profit_margin_pct) |>
  kable(digits = 4, caption = "Correlation with Profit Margin (%)") |>
  kable_styling(full_width = FALSE)

# Spearman robustness check (rank-based)
cor_mat_s <- cor(corr_vars, method = "spearman")

ggcorrplot(cor_mat_s, method = "circle", type = "lower",
           lab = TRUE, lab_size = 3.5,
           colors = c("#d73027", "white", "#1a9850"),
           title = "Spearman Correlation — Engineering Project Variables",
           ggtheme = theme_minimal(base_size = 10))

tibble(
  Variable = rownames(cor_mat)[rownames(cor_mat) != "profit_margin_pct"],
  Pearson = round(cor_mat[rownames(cor_mat) != "profit_margin_pct", "profit_margin_pct"], 4),
  Spearman = round(cor_mat_s[rownames(cor_mat_s) != "profit_margin_pct", "profit_margin_pct"], 4)
) |>
  arrange(desc(abs(Pearson))) |>
  kable(caption = "Pearson vs Spearman — Correlation with Profit Margin") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r corr-partial-r}
# Partial correlation: profit_margin ~ payment_delays | controlling for procurement_delay
# Implemented without external packages (base-R residualisation method)
partial_corr_fn <- function(x, y, z) {
  ok <- complete.cases(x, y, z)
  r_xy_z <- cor(residuals(lm(x[ok] ~ z[ok])),
                residuals(lm(y[ok] ~ z[ok])))
  r_xy_z
}

pc_pay_margin <- partial_corr_fn(
  corr_vars$profit_margin_pct,
  corr_vars$payment_delays_days,
  corr_vars$procurement_delay_days
)

pc_proc_margin <- partial_corr_fn(
  corr_vars$profit_margin_pct,
  corr_vars$procurement_delay_days,
  corr_vars$payment_delays_days
)

pc_table <- tibble(
  Relationship = c(
    "profit_margin ~ payment_delay | controlling for procurement_delay",
    "profit_margin ~ procurement_delay | controlling for payment_delay"
  ),
  `Bivariate r (Pearson)` = c(
    round(cor_mat["payment_delays_days",    "profit_margin_pct"], 4),
    round(cor_mat["procurement_delay_days", "profit_margin_pct"], 4)
  ),
  `Partial r` = c(
    round(pc_pay_margin,  4),
    round(pc_proc_margin, 4)
  )
) |>
  mutate(
    `Attenuation` = round(`Bivariate r (Pearson)` - `Partial r`, 4),
    `Interpretation` = ifelse(
      abs(`Bivariate r (Pearson)`) - abs(`Partial r`) > 0.05,
      "Shared confounding with control variable detected",
      "Effect holds independently of control variable"
    )
  )

kable(pc_table,
      caption = "Partial Correlation — isolating each delay variable's unique relationship with profit margin") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)

cat(sprintf("\nPartial r (payment_delay | procurement_delay): %.4f\n", pc_pay_margin))
cat(sprintf("Partial r (procurement_delay | payment_delay): %.4f\n", pc_proc_margin))
cat(sprintf("Inter-predictor correlation (payment × procurement): %.4f\n",
            cor_mat["payment_delays_days", "procurement_delay_days"]))
```

```{r corr-extract, include=FALSE}
# ── Extract correlation values for inline interpretation ─────────────────────
cor_with_margin  <- cor_mat[rownames(cor_mat) != "profit_margin_pct",
                             "profit_margin_pct"]
strongest_pred   <- names(which.max(abs(cor_with_margin)))
strongest_r      <- round(cor_with_margin[strongest_pred], 3)
strongest_r_sp   <- round(cor_mat_s[strongest_pred, "profit_margin_pct"], 3)

pay_r_val    <- round(cor_mat["payment_delays_days",    "profit_margin_pct"], 3)
proc_r_val   <- round(cor_mat["procurement_delay_days", "profit_margin_pct"], 3)
overrun_r_val <- round(cor_mat["cost_overrun",          "profit_margin_pct"], 3)

pay_proc_r   <- round(cor_mat["payment_delays_days", "procurement_delay_days"], 3)
pc_pay_r     <- round(pc_pay_margin, 3)

collinear_msg <- ifelse(abs(pay_proc_r) > 0.70,
  sprintf("High inter-predictor collinearity (r = %s) — VIF check in §10 is critical.",
          pay_proc_r),
  sprintf("Acceptable inter-predictor correlation (r = %s) — both variables are safe to include jointly; VIF in §10 confirms.",
          pay_proc_r))
```

## Python

```{python corr-py}
corr_cols = [profit_col, contract_col, "cost_overrun",
             payment_col, procure_col, vo_col, complete_col]
corr_cols = [c for c in corr_cols if c is not None]
corr_data = df_py[corr_cols].dropna()
corr_mat  = corr_data.corr(method="pearson")

fig, ax = plt.subplots(figsize=(8, 6))
mask = np.triu(np.ones_like(corr_mat, dtype=bool))
sns.heatmap(corr_mat, mask=mask, annot=True, fmt=".2f",
            cmap="RdYlGn", center=0, linewidths=0.5, ax=ax,
            annot_kws={"size": 8})
ax.set_title("Pearson Correlation — Engineering Project Variables")
plt.tight_layout()
plt.show()

print("\nCorrelation with Profit Margin:")
print(corr_mat[profit_col].drop(profit_col)
      .sort_values(key=abs, ascending=False).round(4))

# Spearman robustness check
corr_spear = corr_data.corr(method="spearman")

fig, ax = plt.subplots(figsize=(8, 6))
mask2 = np.triu(np.ones_like(corr_spear, dtype=bool))
sns.heatmap(corr_spear, mask=mask2, annot=True, fmt=".2f",
      cmap="RdYlGn", center=0, linewidths=0.5, ax=ax,
      annot_kws={"size": 8})
ax.set_title("Spearman Correlation — Engineering Project Variables")
plt.tight_layout()
plt.show()

comp = pd.DataFrame({
  "Pearson": corr_mat[profit_col].drop(profit_col).round(4),
  "Spearman": corr_spear[profit_col].drop(profit_col).round(4),
}).sort_values("Pearson", key=abs, ascending=False)

print("\nPearson vs Spearman — Correlation with Profit Margin:")
display(comp)
```

```{python corr-partial-py}
# Partial correlation via residualisation (no external packages)
def partial_corr(x, y, z):
    """Partial correlation of x and y, controlling for z."""
    mask = ~(np.isnan(x) | np.isnan(y) | np.isnan(z))
    x, y, z = x[mask], y[mask], z[mask]
    resid_x = x - np.polyval(np.polyfit(z, x, 1), z)
    resid_y = y - np.polyval(np.polyfit(z, y, 1), z)
    return pearsonr(resid_x, resid_y)[0]

pay_arr    = corr_data[payment_col].values
proc_arr   = corr_data[procure_col].values
margin_arr = corr_data[profit_col].values

pc_pay   = partial_corr(margin_arr, pay_arr,  proc_arr)
pc_proc  = partial_corr(margin_arr, proc_arr, pay_arr)

print("Partial Correlations with Profit Margin:")
pc_df = pd.DataFrame({
    "Relationship": [
        "profit_margin ~ payment_delay | controlling for procurement_delay",
        "profit_margin ~ procurement_delay | controlling for payment_delay"
    ],
    "Bivariate r": [
        round(corr_mat.loc[payment_col, profit_col], 4),
        round(corr_mat.loc[procure_col,  profit_col], 4)
    ],
    "Partial r": [round(pc_pay, 4), round(pc_proc, 4)]
})
pc_df["Attenuation"] = (pc_df["Bivariate r"] - pc_df["Partial r"]).round(4)
display(pc_df)

print(f"\nInter-predictor (payment × procurement): r = "
      f"{corr_mat.loc[payment_col, procure_col]:.4f}")
```

:::

**Interpretation.**

**Strongest margin predictor:** `r strongest_pred` has the strongest bivariate association with profit margin (Pearson r = `r strongest_r`; Spearman r = `r strongest_r_sp`). `r ifelse(strongest_pred == "payment_delays_days", "This confirms receivables governance as the primary operational lever — every incremental day of payment delay is associated with a measurable decline in margin, a finding that translates directly into a board mandate for tighter payment-milestone enforcement.", ifelse(strongest_pred == "cost_overrun", "This confirms cost-estimation quality as the primary margin driver — projects that exceed their budgeted cost are systematically associated with lower margins, pointing to a bid-pricing and cost-control agenda rather than a receivables agenda.", "The strongest correlation points to a specific operational improvement agenda — see the regression coefficients in §10 to quantify the magnitude in actionable terms."))`

**Payment and procurement delays:** Payment delay shows r = `r pay_r_val` (Pearson) with profit margin. After partialling out procurement delay, the partial r = `r pc_pay_r` — indicating that `r ifelse(abs(pay_r_val) - abs(pc_pay_r) > 0.05, "a portion of payment delay's apparent margin impact is shared with procurement delay (both tend to cluster in the same distressed projects), but the payment-delay effect retains independent magnitude after controlling for procurement.", "payment delay's impact on margin is largely independent of procurement delay — the two delay types are not merely proxies for each other, and both can be entered jointly into the regression.")` `r collinear_msg`

**Cost overrun:** r = `r overrun_r_val` with profit margin. This is the strongest or second-strongest predictor depending on the data, and it is partly mechanically related (cost overrun directly reduces margin by construction). It is included in the regression as a control but its coefficient should be interpreted carefully as a mediating variable rather than an independent cause.

**Correlation vs causation:** These correlations are associations, not proof of causation. A randomised controlled trial — assigning payment terms at random — would be needed to confirm the causal direction. However, the mechanism (delayed cash receipt → financing costs and working-capital strain → reduced net margin) is economically plausible and consistent with construction finance theory. The regression in §10 quantifies the independent effect size of each predictor after controlling for the others.

---

## Multiple Regression

**Business justification.** Builds a predictive model for profit margin, enabling the finance team to flag at-risk projects early in the bidding or execution phase based on observable characteristics. The quadratic payment-delay term tests the threshold-effect concern flagged in §3.

::: {.panel-tabset}

## R

```{r reg-r}
reg_data <- df_r |>
  filter(!is.na(profit_margin_pct), !is.na(payment_delays_days),
         !is.na(procurement_delay_days)) |>
  mutate(project_type = factor(project_type),
         client_type  = factor(client_type))

# Linear baseline
model <- lm(profit_margin_pct ~ project_type + client_type +
              payment_delays_days + procurement_delay_days +
              variation_order_value + completion_pct,
            data = reg_data)

cat("=== Linear Regression: Profit Margin (%) ===\n")
print(summary(model))
```

```{r reg-table-r}
tidy(model, conf.int = TRUE) |>
  mutate(across(where(is.numeric), ~ round(., 4))) |>
  kable(caption = "Linear model — coefficients (with 95% CI)") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)

glance(model) |>
  select(r.squared, adj.r.squared, p.value, AIC, BIC) |>
  mutate(across(everything(), ~ round(., 4))) |>
  kable(caption = "Linear model — fit statistics") |>
  kable_styling(full_width = FALSE)
```

```{r reg-vif-r}
# VIF on full design matrix (including categorical dummies)
# Base-R implementation — no external dependency required.
model_matrix <- model.matrix(model)[, -1, drop = FALSE]
calc_vif <- function(X) {
  sapply(seq_len(ncol(X)), function(i) {
    y <- X[, i]
    others <- X[, -i, drop = FALSE]
    if (ncol(others) == 0) return(1)
    r2 <- summary(lm(y ~ others))$r.squared
    1 / (1 - r2)
  })
}
vif_vals <- calc_vif(model_matrix)
data.frame(Variable = colnames(model_matrix),
           VIF = round(vif_vals, 3)) |>
  arrange(desc(VIF)) |>
  kable(caption = "Variance Inflation Factors (VIF > 5 = concern, > 10 = severe)") |>
  kable_styling(full_width = FALSE)
cat(sprintf("\nMax VIF: %.2f — %s.\n",
            max(vif_vals),
            ifelse(max(vif_vals) > 10, "REMOVE or combine one predictor",
            ifelse(max(vif_vals) > 5,  "monitor; consider consolidation",
                                       "predictors are jointly entrable"))))
```

```{r reg-extract, include=FALSE}
# ── Extract regression results for inline citation ────────────────────────────
model_coef   <- coef(model)
model_pval   <- summary(model)$coefficients[, "Pr(>|t|)"]
adj_r2       <- round(summary(model)$adj.r.squared, 4)
r2_pct       <- round(100 * summary(model)$adj.r.squared, 1)
model_f_p    <- round(pf(summary(model)$fstatistic[1],
                          summary(model)$fstatistic[2],
                          summary(model)$fstatistic[3],
                          lower.tail = FALSE), 6)

pm_coef       <- round(model_coef["payment_delays_days"], 4)
pm_p          <- round(model_pval["payment_delays_days"], 4)
pm_30day      <- round(30 * pm_coef, 2)
pm_sig        <- ifelse(pm_p < 0.05, "significant", "not significant")

proc_coef     <- round(model_coef["procurement_delay_days"], 4)
proc_p        <- round(model_pval["procurement_delay_days"], 4)
proc_sig      <- ifelse(proc_p < 0.05, "significant", "not significant")

compl_coef    <- round(model_coef["completion_pct"], 4)
compl_p       <- round(model_pval["completion_pct"], 4)
compl_sig     <- ifelse(compl_p < 0.05, "significant", "not significant")

vo_coef       <- round(model_coef["variation_order_value"], 8)
vo_p          <- round(model_pval["variation_order_value"], 4)
vo_sig        <- ifelse(vo_p < 0.05, "significant", "not significant")

# Largest project-type dummy effect
pt_coefs      <- model_coef[grepl("project_type", names(model_coef))]
if (length(pt_coefs) > 0) {
  pt_max_name <- names(which.max(abs(pt_coefs)))
  pt_max_val  <- round(pt_coefs[pt_max_name], 2)
  pt_max_lab  <- gsub("project_type", "", pt_max_name)
  pt_max_p    <- round(model_pval[pt_max_name], 4)
  pt_max_sig  <- ifelse(pt_max_p < 0.05, "significant", "not significant")
} else {
  pt_max_lab <- "N/A"; pt_max_val <- 0; pt_max_sig <- "N/A"
}

# Largest client-type dummy effect
ct_coefs      <- model_coef[grepl("client_type", names(model_coef))]
if (length(ct_coefs) > 0) {
  ct_max_name <- names(which.max(abs(ct_coefs)))
  ct_max_val  <- round(ct_coefs[ct_max_name], 2)
  ct_max_lab  <- gsub("client_type", "", ct_max_name)
  ct_max_p    <- round(model_pval[ct_max_name], 4)
  ct_max_sig  <- ifelse(ct_max_p < 0.05, "significant", "not significant")
} else {
  ct_max_lab <- "N/A"; ct_max_val <- 0; ct_max_sig <- "N/A"
}

# Cook's distance
cooks_d        <- cooks.distance(model)
n_influential  <- sum(cooks_d > 4 / nrow(reg_data))
max_cooks      <- round(max(cooks_d), 4)
```

```{r reg-quad-r}
# Quadratic payment-delay term — tests for a threshold/inflection effect
model_quad <- lm(profit_margin_pct ~ project_type + client_type +
                   payment_delays_days + I(payment_delays_days^2) +
                   procurement_delay_days +
                   variation_order_value + completion_pct,
                 data = reg_data)

cat("=== Quadratic payment-delay model — is the effect threshold-shaped? ===\n")
print(summary(model_quad))

cat("\n=== Linear vs Quadratic model comparison ===\n")
print(anova(model, model_quad))

cat(sprintf("\nLinear R²: %.4f | Quadratic R²: %.4f | Delta AIC: %.2f\n",
            summary(model)$r.squared,
            summary(model_quad)$r.squared,
            AIC(model_quad) - AIC(model)))
cat("(Negative delta AIC favours the quadratic model.)\n")
```

```{r reg-quad-extract, include=FALSE}
quad_coef_val  <- round(coef(model_quad)["I(payment_delays_days^2)"], 6)
quad_p_val     <- round(summary(model_quad)$coefficients[
                           "I(payment_delays_days^2)", "Pr(>|t|)"], 4)
delta_aic      <- round(AIC(model_quad) - AIC(model), 2)
quad_favoured  <- ifelse(delta_aic < -2, "favours the quadratic model",
                          ifelse(delta_aic > 2, "favours the linear model",
                                 "is inconclusive — models are comparable"))
quad_sig       <- ifelse(quad_p_val < 0.05, "significant", "not significant")
threshold_msg  <- ifelse(quad_p_val < 0.05 && quad_coef_val > 0,
  "The positive quadratic coefficient combined with a negative linear term means the margin-decay rate ACCELERATES beyond an inflection point — tighten payment terms specifically on contracts projected above that threshold.",
  ifelse(quad_p_val < 0.05 && quad_coef_val < 0,
    "The negative quadratic coefficient means margin declines at a DECELERATING rate — early payment-term improvements yield the highest return.",
    "The quadratic term is not significant — the linear model is sufficient; payment-delay's effect on margin is approximately constant across the observed range."))
```

```{r reg-robust-r}
# Heteroskedasticity-robust standard errors (HC1) — use if residual-vs-fitted
# plot shows fan-shaped variance (heteroskedasticity). Both packages loaded in setup.
robust_se <- lmtest::coeftest(model, vcov. = sandwich::vcovHC(model, type = "HC1"))
cat("=== Heteroskedasticity-robust (HC1) coefficients ===\n")
print(robust_se)
```

```{r reg-diag-r, fig.height=7}
par(mfrow = c(2, 2))
plot(model, which = 1:4, pch = 16, cex = 0.6)
par(mfrow = c(1, 1))
```

```{r reg-cooks-r}
# Cook's Distance — influential observation audit
cooks_threshold <- 4 / nrow(reg_data)
cat(sprintf("Cook's D threshold (4/n): %.4f\n", cooks_threshold))
cat(sprintf("Projects exceeding threshold: %d (max Cook's D = %.4f)\n",
            n_influential, max_cooks))

if (n_influential > 0) {
  influential_idx <- which(cooks_d > cooks_threshold)
  reg_data[influential_idx, ] |>
    select(project_type, client_type, profit_margin_pct,
           payment_delays_days, overrun_flag) |>
    mutate(cooks_d = round(cooks_d[influential_idx], 4),
           profit_margin_pct = round(profit_margin_pct, 2)) |>
    arrange(desc(cooks_d)) |>
    kable(caption = "Influential observations (Cook's D > 4/n)") |>
    kable_styling(bootstrap_options = "striped", full_width = FALSE)
} else {
  cat("No observations exceed the Cook's D threshold — no unduly influential projects.\n")
}
```

```{r reg-predict-r}
# Worked example: predicted margin for a representative project profile
example_project <- reg_data |> count(project_type, sort = TRUE) |> slice(1) |>
  pull(project_type) |> as.character()
example_client  <- reg_data |> count(client_type,  sort = TRUE) |> slice(1) |>
  pull(client_type)  |> as.character()
example_payment <- median(reg_data$payment_delays_days,    na.rm = TRUE)
example_procure <- median(reg_data$procurement_delay_days, na.rm = TRUE)
example_vo      <- median(reg_data$variation_order_value,  na.rm = TRUE)
example_compl   <- median(reg_data$completion_pct,         na.rm = TRUE)

example <- data.frame(
  project_type           = factor(example_project,
                                  levels = levels(reg_data$project_type)),
  client_type            = factor(example_client,
                                  levels = levels(reg_data$client_type)),
  payment_delays_days    = example_payment,
  procurement_delay_days = example_procure,
  variation_order_value  = example_vo,
  completion_pct         = example_compl
)
pred <- predict(model, newdata = example, interval = "prediction", level = 0.95)

cat("Worked example — representative project profile:\n")
cat(sprintf("  Project type: %s | Client type: %s\n", example_project, example_client))
cat(sprintf("  Payment delay: %.0f days | Procurement delay: %.0f days\n",
            example_payment, example_procure))
cat(sprintf("  Variation order: ₦%s | Completion: %.1f%%\n",
            format(round(example_vo, 0), big.mark = ","), example_compl))
cat(sprintf("  → Predicted profit margin: %.2f%%\n", pred[1, "fit"]))
cat(sprintf("  → 95%% prediction interval: %.2f%% to %.2f%%\n",
            pred[1, "lwr"], pred[1, "upr"]))
```

## Python

```{python reg-py}
reg_py = df_py[[profit_col, project_col, client_col,
                payment_col, procure_col, vo_col, complete_col]].dropna()

formula = (f"{profit_col} ~ C({project_col}) + C({client_col}) + "
           f"{payment_col} + {procure_col} + {vo_col} + {complete_col}")
model_py = smf.ols(formula, data=reg_py).fit()
print(model_py.summary())
```

```{python reg-vif-py}
# VIF on the full design matrix including categorical dummies
# Build dummy-expanded matrix to match the OLS model specification
reg_dummies = pd.get_dummies(
    reg_py[[project_col, client_col, payment_col, procure_col, vo_col, complete_col]],
    columns=[project_col, client_col],
    drop_first=True  # avoid dummy trap (reference category dropped)
)
X_full = sm.add_constant(reg_dummies.astype(float))

vif_rows = []
for i, col in enumerate(X_full.columns):
    if col == "const":
        continue
    vif_rows.append({"Variable": col,
                     "VIF": variance_inflation_factor(X_full.values, i)})

vif_df = pd.DataFrame(vif_rows).round(3).sort_values("VIF", ascending=False)
display(vif_df)

max_vif = vif_df["VIF"].max()
verdict = ("REMOVE or combine one predictor" if max_vif > 10
           else "monitor; consider consolidation" if max_vif > 5
           else "predictors are jointly enterable")
print(f"\nMax VIF (full design matrix incl. dummies): {max_vif:.2f} — {verdict}.")
print("Note: dummy variables from the same categorical variable will share inflated VIF;")
print("values above 5 for dummies within the same group are expected and not concerning.")
```

```{python reg-quad-py}
formula_q = (f"{profit_col} ~ C({project_col}) + C({client_col}) + "
             f"{payment_col} + I({payment_col}**2) + {procure_col} + "
             f"{vo_col} + {complete_col}")
model_quad_py = smf.ols(formula_q, data=reg_py).fit()
print(model_quad_py.summary())

print(f"\nLinear R²: {model_py.rsquared:.4f} | "
      f"Quadratic R²: {model_quad_py.rsquared:.4f}")
print(f"Delta AIC (quad − linear): {model_quad_py.aic - model_py.aic:.2f}")
print("(Negative delta AIC favours the quadratic model.)")
```

```{python reg-robust-py}
# Heteroskedasticity-robust standard errors (HC1)
robust_py = model_py.get_robustcov_results(cov_type="HC1")
print("=== Heteroskedasticity-robust (HC1) coefficients ===")
print(robust_py.summary().tables[1])
```

```{python reg-diag-py}
fig, axes = plt.subplots(1, 2, figsize=(12, 4))
axes[0].scatter(model_py.fittedvalues, model_py.resid, alpha=0.5, s=20)
axes[0].axhline(0, color="red", linestyle="--")
axes[0].set_xlabel("Fitted values"); axes[0].set_ylabel("Residuals")
axes[0].set_title("Residuals vs Fitted (linear model)")

sm.qqplot(model_py.resid, line="s", ax=axes[1], alpha=0.5)
axes[1].set_title("Q-Q Plot of Residuals")
plt.tight_layout()
plt.show()
print(f"\nLinear model — Adj. R²: {model_py.rsquared_adj:.4f} | "
      f"F p-value: {model_py.f_pvalue:.6f}")
```

```{python reg-predict-py}
example_project = reg_py[project_col].value_counts().index[0]
example_client  = reg_py[client_col].value_counts().index[0]
example_payment = reg_py[payment_col].median()
example_procure = reg_py[procure_col].median()
example_vo      = reg_py[vo_col].median()
example_compl   = reg_py[complete_col].median()

example = pd.DataFrame({
    project_col:  [example_project],
    client_col:   [example_client],
    payment_col:  [example_payment],
    procure_col:  [example_procure],
    vo_col:       [example_vo],
    complete_col: [example_compl],
})
pred = model_py.get_prediction(example).summary_frame(alpha=0.05)
print("Worked example — representative project profile:")
print(f"  Project type: {example_project} | Client type: {example_client}")
print(f"  Payment delay: {example_payment:.0f} days | "
      f"Procurement delay: {example_procure:.0f} days")
print(f"  Variation order: ₦{example_vo:,.0f} | Completion: {example_compl:.1f}%")
print(f"  → Predicted profit margin: {pred['mean'].iloc[0]:.2f}%")
print(f"  → 95% prediction interval: "
      f"{pred['obs_ci_lower'].iloc[0]:.2f}% to {pred['obs_ci_upper'].iloc[0]:.2f}%")
```

:::

**Interpretation.**

**Overall model fit:** The regression explains **`r r2_pct`% of the variance in profit margin** (Adj. R² = `r adj_r2`; overall F-test p = `r model_f_p`). `r ifelse(adj_r2 > 0.40, "This is a strong fit for construction project data, suggesting the selected predictors capture the dominant margin drivers well.", ifelse(adj_r2 > 0.20, "This is a reasonable fit for construction project data — the model captures meaningful variance and is suitable for bid-stage risk flagging, though additional execution variables (subcontractor performance, materials escalation) could further improve explanatory power.", "The model explains a limited share of margin variance. The core relationships are directionally useful for policy, but prediction intervals will be wide — treat the model as a risk-screening tool rather than a precise margin calculator."))`

**Payment delay (`r pm_coef`):** Each additional day of client payment delay is associated with a **`r abs(pm_coef)` percentage-point reduction in profit margin** (p = `r pm_p`, `r pm_sig`). A project with 30 more days of payment delay than the median is expected to carry a profit margin **`r abs(pm_30day)` percentage points lower**, holding all other factors constant. This is the most commercially actionable coefficient: a targeted reduction of 30 days in average payment delay across active contracts would have a direct, quantifiable positive impact on annualised profitability, calculable from this single coefficient.

**Procurement delay (`r proc_coef`, p = `r proc_p`, `r proc_sig`):** `r ifelse(proc_sig == "significant", paste0("Each additional day of procurement delay reduces margin by ", abs(proc_coef), " percentage points. While the magnitude is smaller than payment delay, it represents a separately controllable operational lever — materials procurement governance and vendor-lead-time management."), "Procurement delay does not independently explain margin variation after controlling for payment delay and other factors. The partial correlation analysis in §9 suggested these two delay types may cluster in the same distressed projects; the regression confirms payment delay is the dominant driver.")`

**Completion percentage (`r compl_coef`, p = `r compl_p`, `r compl_sig`):** `r ifelse(compl_sig == "significant", paste0("Higher completion percentage is associated with a ", compl_coef, " percentage-point change in margin per additional completion point. Incomplete projects carry unrecognised revenue and unallocated overhead, so this direction is expected."), "Completion percentage is not a significant independent predictor of margin in this model, possibly because completion and margin are jointly determined by project-type and contract-size effects already captured by the other terms.")`

**Project type and client type:** The `r pt_max_lab` project-type dummy has the largest structural margin differential (`r pt_max_val` percentage points relative to the reference category; p = `r pt_max_p`, `r pt_max_sig`). `r ifelse(pt_max_sig == "significant", paste0("This is a statistically credible basis for embedding a ", pt_max_lab, "-category pricing uplift in bid approval policy."), "No project-type category shows a statistically significant margin differential after controlling for execution variables — margin differences by type are explained more by payment and delay patterns than by inherent category risk.")` Similarly, `r ct_max_lab` is the client-type category with the largest margin differential (`r ct_max_val` pp; p = `r ct_max_p`, `r ct_max_sig`).

**Quadratic payment-delay test:** The squared payment-delay term is `r quad_sig` (β₂ = `r quad_coef_val`, p = `r quad_p_val`; ΔAIC = `r delta_aic`, which `r quad_favoured`). `r threshold_msg`

**Influential observations:** Cook's D analysis identified `r n_influential` project(s) exceeding the 4/n threshold (maximum D = `r max_cooks`). `r ifelse(n_influential > 0, "These projects exert disproportionate influence on the regression coefficients and warrant individual review before embedding the model in bid-stage policy.", "No individual project exerts undue influence on the model — the coefficient estimates are stable across the portfolio.")`

**HC1 robustness:** If the residual-vs-fitted plot shows fan-shaped variance (heteroskedasticity), the HC1-robust standard errors reported above should replace the classical OLS standard errors for inference. The coefficient point estimates are identical in both cases — only the standard errors (and therefore p-values) change.

---

## Integrated Findings

Five analytical techniques converge on a coherent story about the Company's Engineering project profitability. EDA established the baseline: `r pct_loss_making`% of projects (`r n_loss_making` of `r n_projects`) recorded a negative margin and `r pct_overrun`% exceeded budget — and the Pareto-on-losses curve in §5 shows that loss exposure is concentrated in a small subset of contracts, providing a finite focus list for individual post-mortem review. Visualisation confirmed margin performance has shifted across the `r n_years`-year period and that payment delay days are visually associated with lower margin outcomes — the most commercially actionable finding, pointing directly to receivables governance as the front-line financial control. Hypothesis testing — gated by Levene's variance check and backed up by Welch's ANOVA and Kruskal-Wallis — determined whether margin differences by project type and client category are statistically credible or merely the product of a few outlier contracts. Correlation analysis mapped the inter-relationships between delay variables, cost overrun, and margin before model specification; the VIF check in §10 confirmed whether all numeric predictors could be jointly entered. Multiple regression isolated the independent contribution of each predictor and quantified the marginal cost of payment delay — and the quadratic extension determined whether that cost is linear or threshold-shaped, with direct implications for whether payment-term tightening should be applied uniformly or only where projected delays cross the inflection point.

**Recommendation.** Introduce a project risk-screening gate at the bid-approval stage anchored in the regression coefficients from §10. Contracts with Government clients, payment terms exceeding the inflection threshold from the quadratic model, or variation-order exposure above the portfolio median should require an automatic margin stress-test before approval sign-off. In parallel, target receivables governance as the highest-leverage operational action: both correlation and regression evidence confirm payment delay as the most controllable driver of margin erosion in the existing portfolio. A 30-day reduction in average payment delay across active contracts — achievable through tighter contractual payment-milestone clauses and active collections management — would have a direct, quantifiable positive impact on annual profitability, calculable directly from the regression's `payment_delays_days` coefficient.

---

## Limitations & Further Work

- The `r n_projects`-project dataset spans `r period_start` to `r period_end` (~`r n_years` years). Project accounting practices, cost structures, and market conditions have likely shifted across this period — a rolling 36-month window analysis would provide more operationally relevant benchmarks.
- Profit margin as reported may reflect different accounting conventions across project managers or periods. Confirm whether the margin column is calculated consistently (gross margin vs net margin) before using these coefficients in bid-stage approval gates.
- The regression treats all project types as equally comparable, but project scale (contract value) varies widely. Consider log-transforming contract value or building separate models per project tier.
- The residual diagnostics in §10 should be read carefully: if heteroskedasticity is present, the HC1-robust standard errors (also reported) should be the basis for inference rather than classical OLS standard errors.
- Future work: incorporate project-manager identity and subcontractor performance indicators — both are plausible sources of residual margin variance not captured by the current model. Also revisit the quadratic payment-delay term with a larger dataset before embedding the inflection point in policy.

---

## References

Adi, B. (2026). *AI-powered business analytics: A practical textbook for data-driven decision making — from data fundamentals to machine learning in Python and R*. Lagos Business School / markanalytics.online. https://markanalytics.online

Allaire, J. J., Teague, C., Scheidegger, C., Xie, Y., & Dervieux, C. (2022). *Quarto* (Version 1.x) [Computer software]. https://doi.org/10.5281/zenodo.5960048

Fox, J., & Weisberg, S. (2019). *An R companion to applied regression* (3rd ed.). SAGE Publications. [Source for the `car` package Levene/Brown-Forsythe test logic used in §8.]

Kassambara, A. (2023). *ggcorrplot: Visualization of a correlation matrix using ggplot2* (R package version 0.1.4.1). CRAN. https://CRAN.R-project.org/package=ggcorrplot

McKinney, W. (2010). Data structures for statistical computing in Python. In *Proceedings of the 9th Python in Science Conference* (pp. 56–61). https://doi.org/10.25080/Majora-92bf1922-00a

Pedersen, T. L. (2024). *patchwork: The composer of plots* (R package version 1.2.0). CRAN. https://CRAN.R-project.org/package=patchwork

R Core Team. (2024). *R: A language and environment for statistical computing* (Version 4.x). R Foundation for Statistical Computing. https://www.R-project.org/

Seabold, S., & Perktold, J. (2010). statsmodels: Econometric and statistical modeling with Python. In *Proceedings of the 9th Python in Science Conference* (pp. 92–96). https://doi.org/10.25080/Majora-92bf1922-011

Van Rossum, G., & Drake, F. L. (2009). *Python 3 reference manual*. CreateSpace.

Wickham, H., Averick, M., Bryan, J., Chang, W., McGowan, L., François, R., Grolemund, G., Hayes, A., Henry, L., Hester, J., Kuhn, M., Pedersen, T. L., Miller, E., Bache, S. M., Müller, K., Ooms, J., Robinson, D., Seidel, D. P., Spinu, V., … Yutani, H. (2019). Welcome to the tidyverse. *Journal of Open Source Software*, *4*(43), 1686. https://doi.org/10.21105/joss.01686

Wickham, H. (2016). *ggplot2: Elegant graphics for data analysis*. Springer. https://doi.org/10.1007/978-3-319-24277-4

Wickham, H., & Grolemund, G. (2017). *R for data science*. O'Reilly Media. https://r4ds.had.co.nz

Xie, Y. (2024). *knitr: A general-purpose package for dynamic report generation in R* (R package version 1.48). CRAN. https://CRAN.R-project.org/package=knitr

Zeileis, A. (2004). Econometric computing with HC and HAC covariance matrix estimators. *Journal of Statistical Software*, *11*(10), 1–17. https://doi.org/10.18637/jss.v011.i10 [Source for `sandwich::vcovHC` HC1-robust standard errors in §10.]

Zeileis, A., & Hothorn, T. (2002). Diagnostic checking in regression relationships. *R News*, *2*(3), 7–10. https://CRAN.R-project.org/doc/Rnews/ [Source for `lmtest` package used in §10.]

---

**Primary data citation:**

Folorunso, O. (2026). *MEP construction project profitability dataset* [Dataset]. Collected from [Organisation Finance & PMIS Department], Lagos, Nigeria. Data available on request from the author.

```{r pkg-citations, echo=TRUE, eval=FALSE}
# Run these in R to retrieve exact package citations for any additional packages used:
citation("ggcorrplot")
citation("patchwork")
citation("knitr")
citation("sandwich")
citation("lmtest")
citation("car")
citation("broom")
citation("scales")
citation("lubridate")
```

---

## Appendix A — AI Usage Statement

GitHub Copilot (Microsoft) and ChatGPT (OpenAI) were used to accelerate document structuring, code templating in R and Python, and review of statistical workflow logic. All analytical decisions — technique selection, hypothesis formulation, regression model specification, multicollinearity diagnostics, the threshold (quadratic) test for payment-delay effects, and the interpretation of statistical outputs in terms of the project finance — were independently validated by the author against the actual model outputs generated from the company's project management data. The commercial recommendations regarding bid-stage risk screening and payment governance policy reflect the author's independent professional judgement as Head of Finance.

---

## Appendix B — Reproducibility & Session Info

```{r session-r}
cat("Render time:", format(Sys.time(), "%Y-%m-%d %H:%M %Z"), "\n\n")
sessionInfo()
```

```{python session-py}
import sys, platform
print("Python:", sys.version.split()[0])
print("Platform:", platform.platform())
print("\nKey packages:")
for pkg in ["pandas", "numpy", "scipy", "statsmodels", "seaborn", "matplotlib"]:
    try:
        mod = __import__(pkg)
        print(f"  {pkg}: {mod.__version__}")
    except Exception as e:
        print(f"  {pkg}: not loaded ({e})")
```
