---
title: "Project Profitability Analytics: Drivers of Profit Margin in  Construction Projects"
author: "Oladayo Folorunso"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
execute:
  warning: false
  message: false
---

```{r setup, include=FALSE}
if (requireNamespace("reticulate", quietly = TRUE)) {
  venv_python <- "/Users/user/Documents/LandMan/UA High/LBA -EMBA -Nomso/.venv/bin/python"
  if (file.exists(venv_python)) {
    reticulate::use_python(venv_python, required = FALSE)
  }
}

pkgs <- c("readr", "dplyr", "tidyr", "tibble", "purrr", "forcats", "ggplot2",
          "lubridate", "scales", "ggcorrplot", "patchwork", "broom", "knitr")
invisible(lapply(pkgs, library, character.only = TRUE))
if (!requireNamespace("kableExtra", quietly = TRUE)) {
  kable_styling <- function(x, ...) x
} else {
  library(kableExtra)
}
clean_names <- function(df) {
  nms <- names(df)
  nms <- tolower(nms)
  nms <- gsub("[^a-z0-9]+", "_", nms)
  nms <- gsub("_+", "_", nms)
  nms <- gsub("^_|_$", "", nms)
  names(df) <- nms
  df
}
if (requireNamespace("janitor", quietly = TRUE)) {
  clean_names <- janitor::clean_names
}
knitr::opts_chunk$set(fig.width = 9, fig.height = 5, dpi = 150)
```

```{python setup-py, include=FALSE}
import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
import matplotlib.ticker as mticker
import seaborn as sns
from scipy import stats
from scipy.stats import f_oneway, kruskal, ttest_ind, pearsonr
import statsmodels.formula.api as smf
import statsmodels.api as sm
from statsmodels.stats.multicomp import pairwise_tukeyhsd
from IPython.display import display
sns.set_theme(style="whitegrid", palette="muted")
plt.rcParams["figure.dpi"] = 150
```

```{r load-r, include=FALSE}
# ⚠ Column names in raw CSV have leading/trailing whitespace —
# janitor::clean_names() resolves all of these automatically.
df_r <- read_csv("data/MEP_construction_company_projs.csv",
                 col_types = cols(.default = col_character())) |>
  clean_names() |>    # strips whitespace, lowercases, converts spaces → underscores
  mutate(
    month               = my(month),           # "Apr-16" → Date
    contract_value      = as.numeric(gsub("[^0-9.]", "", contract_value)),
    budgeted_cost       = as.numeric(gsub("[^0-9.]", "", budgeted_cost)),
    actual_cost         = as.numeric(gsub("[^0-9.]", "", actual_cost)),
    profit_margin_pct   = as.numeric(gsub("[^0-9.\\-]", "", profit_margin)),
    payment_delays_days = as.numeric(payment_delays_days),
    variation_order_value    = as.numeric(gsub("[^0-9.\\-]", "", variation_order_value)),
    procurement_delay_days   = as.numeric(procurement_delay_days),
    completion_pct           = as.numeric(gsub("[^0-9.]", "", completion))
  ) |>
  filter(!is.na(profit_margin_pct))

# Derived: cost overrun flag
df_r <- df_r |>
  mutate(cost_overrun = actual_cost - budgeted_cost,
         overrun_flag = if_else(cost_overrun > 0, "Over Budget", "On/Under Budget"))
```

```{python load-py, include=FALSE}
df_py = pd.read_csv("data/MEP_construction_company_projs.csv", dtype=str)

# ⚠ Strip leading/trailing whitespace from all column names
df_py.columns = df_py.columns.str.strip()

# Normalise to snake_case
df_py.columns = (df_py.columns.str.lower()
                 .str.replace(r"[^\w]", "_", regex=True)
                 .str.replace(r"_+", "_", regex=True)
                 .str.strip("_"))

# Parse month
df_py["month"] = pd.to_datetime(df_py["month"].str.strip(), format="%b-%y", errors="coerce")

# Numeric columns
num_cols = ["contract_value", "budgeted_cost", "actual_cost",
            "profit_margin", "payment_delays_days",
            "variation_order_value", "procurement_delay_days", "completion"]
for col in num_cols:
    # match against clean column name
    match = [c for c in df_py.columns if col.replace("_", "") in c.replace("_", "")]
    if match:
        df_py[match[0]] = pd.to_numeric(
            df_py[match[0]].str.replace(r"[^\d.\-]", "", regex=True), errors="coerce")

# Re-identify key columns by partial match for safety
def find_col(df, keyword):
    matches = [c for c in df.columns if keyword in c]
    return matches[0] if matches else None

profit_col   = find_col(df_py, "profit_margin")
payment_col  = find_col(df_py, "payment_delay")
procure_col  = find_col(df_py, "procurement_delay")
actual_col   = find_col(df_py, "actual_cost")
budget_col   = find_col(df_py, "budgeted_cost")
contract_col = find_col(df_py, "contract_value")
project_col  = find_col(df_py, "project_type")
client_col   = find_col(df_py, "client_type")
complete_col = find_col(df_py, "completion")
vo_col       = find_col(df_py, "variation_order")

df_py = df_py.dropna(subset=[profit_col])
df_py["cost_overrun"] = df_py[actual_col] - df_py[budget_col]
df_py["overrun_flag"] = np.where(df_py["cost_overrun"] > 0, "Over Budget", "On/Under Budget")
```

---

## 1. Executive Summary

**Business problem:** Profit margin variability is the central financial risk for the construction project delivery. Understanding which project characteristics — type, client category, payment behaviour, procurement delays — drive margin erosion enables the finance function to price contracts more accurately and implement early-warning controls during execution.

**Data collected:** 106 construction projects executed between April 2016 and January 2026, drawn from the company's project management information system. Metrics include contract value, budgeted and actual cost, profit margin (%), payment delays, variation orders, procurement delays, and project completion percentage.

**Key findings:**

1. Margin performance differs across project and client categories, confirming that portfolio mix is a major profitability driver.
2. Delay variables are negatively associated with margin strength, reinforcing the financing and execution risk pathway.
3. Regression results show that controllable execution and cash-flow variables explain a meaningful share of margin variation.

**Recommendation:** Introduce delay-sensitive pricing guardrails and tighter procurement/payment controls for high-risk project profiles. This should improve median margin performance and reduce loss-making project incidence.

---

## 2. Professional Disclosure

**Job title:** Head of Finance (CFO-designate)  
**Organisation sector:** Construction and project delivery

### Technique 1 — Exploratory Data Analysis
As Head of Finance overseeing a 106-project construction portfolio spanning April 2016 to January 2026, my most pressing strategic concern is margin volatility: why do some projects deliver healthy returns while others erode the company's financial position? Before any formal test, I need to understand the *shape* of that margin distribution — what percentage of projects are loss-making, which project type clusters exhibit the widest variance, and whether the most extreme negative margins are genuine contractual losses or accounting artefacts such as variation orders not yet captured in the cost system. EDA establishes this baseline and determines whether the distributional characteristics support parametric ANOVA and regression, or whether non-parametric alternatives are required. This mirrors the quarterly margin review I prepare for the MD, where the opening chart always presents the distribution of project-level performance before any cause-and-effect analysis.

**Alternative considered:** Starting with a regression model on the raw data to identify margin drivers directly. Rejected because undetected data quality issues — inconsistent margin calculations across project managers, missing variation order entries, or incomplete cost records for in-progress projects — would silently bias coefficients and produce recommendations based on artefactual rather than genuine performance patterns.

**Limitation:** EDA is descriptive only. A project type that appears to underperform on average may still not do so significantly once the variance and sample size are properly weighted in a formal test. ANOVA and regression are required before making project portfolio or bid-pricing decisions based on observed margin differences.

### Technique 2 — Data Visualization
Margin analytics in construction project finance serve a dual purpose: internal project management and external commercial strategy. The time-series chart of annual average profit margin by project type (2016–2026) answers a board-level question — is our portfolio profitability improving or deteriorating, and are particular project segments driving a structural decline? The scatter plot of payment delay days against profit margin translates the most operationally controllable risk variable — receivables ageing — into a visual argument for tighter collection governance: a clearly negative relationship provides the finance function with board-presentable evidence that payment behaviour directly affects company profitability. The cost overrun rate chart by project type provides both the procurement team and the MD with the input needed for risk-adjusted contracting and contingency provisioning.

**Alternative considered:** Presenting only the regression output without supporting visualizations. Rejected because regression coefficients require contextual interpretation, and non-technical stakeholders — project managers, the MD, and the board — rely on visual communication to understand commercial implications. Charts also reveal non-linear patterns that the linear regression model may not capture.

**Limitation:** A 10-year window introduces potential confounding from macroeconomic cycles, construction material cost inflation post-2022, and changes in company project intake policy. Time trends in the visualization should be interpreted with awareness that the market environment has shifted materially across this period, and margin benchmarks from 2016–2020 may not be applicable to current bid pricing.

### Technique 3 — Hypothesis Testing
A recurrent finance-team debate is whether certain project categories or client types are structurally less profitable — and whether that difference is a reliable basis for adjusting bid strategy and risk provisioning, or merely a noise artefact from a small number of outlier projects. One-way ANOVA on profit margin across project types (Commercial, Residential, Industrial, Infrastructure) and across client types (Government, Private, Multinational) converts this qualitative observation into a defensible statistical claim. If ANOVA confirms a significant difference — for example, Government-client projects delivering systematically lower margin than Private-sector — that result supports a formal pricing adjustment and risk-uplift policy for public-sector bids: a recommendation that can be taken to the MD with a p-value, not just an opinion.

**Alternative considered:** A simple comparison of highest- versus lowest-performing project type means. Rejected because this approach ignores variance and sample size heterogeneity, allowing extreme outliers to dominate the narrative. Formal ANOVA with Tukey HSD post-hoc testing controls for multiple comparisons and provides a statistically sound basis for identifying which specific type pairings are genuinely different rather than artefactually divergent.

**Limitation:** Some project type or client type subgroups contain fewer than 20 observations. For these smaller groups, the Tukey HSD test has limited statistical power, and the absence of a significant result should not be taken as proof of equality. The non-parametric Kruskal-Wallis test runs as a robustness check.

### Technique 4 — Correlation Analysis
The five candidate predictors for the profit margin regression model — payment delay days, procurement delay days, cost overrun, variation order value, and completion percentage — may be partly collinear. Projects with long payment delays may also experience procurement delays if cash-flow constraints defer materials procurement, inflating the apparent effect of both variables simultaneously if they are entered into the regression without screening. Correlation analysis maps these inter-variable relationships before model specification, identifying pairs with |r| > 0.70 that require careful handling to avoid unstable coefficient estimates. It also confirms the direction and magnitude of each variable's relationship with profit margin — establishing which operational levers are empirically linked to margin outcomes before any model is built.

**Alternative considered:** Including all five predictors in the regression without prior screening. Rejected because high multicollinearity between payment and procurement delay variables could produce coefficient sign reversals or dramatically inflated standard errors, making the model practically unreliable for the purpose of identifying margin improvement levers.

**Limitation:** Pearson correlation captures only linear associations. If the relationship between payment delay and margin is threshold-based — for example, margin is minimally affected for delays below 60 days but drops sharply beyond that — the correlation coefficient understates the true relationship. Scatterplots in the visualization section provide complementary evidence on functional form.

### Technique 5 — Multiple Regression
The central analytical objective of this project is a project margin prediction model: given a contract's type, client category, payment history, procurement lead time, and variation order exposure, what is the expected profit margin? As Head of Finance, this model enables me to flag high-risk contracts at the bid stage before resources are committed, build differentiated contingency provisions in the budget plan, and quantify the margin cost of accepting unfavourable payment terms. The regression coefficient on `payment_delays_days` directly answers: "What is the marginal cost — in percentage-point margin terms — of accepting a contract with 30 additional days of payment delay?" This is a number with immediate implications for contract negotiation and bid pricing, translating an analytical output into a commercial policy tool.

**Alternative considered:** Building a margin classification model (profitable vs. loss-making) using logistic regression rather than a continuous margin prediction. Rejected because the business question is not binary — it concerns the *degree* of margin performance — and a continuous OLS model provides more granular information for bid pricing, contingency quantification, and financial planning than a simple binary classifier.

**Limitation:** The regression assumes that the marginal effect of each predictor is constant across the 10-year observation window. Given the construction materials cost shock of 2022–2023 and exchange rate pressures on imported  equipment, older project records may reduce the contemporary relevance of the estimated coefficients. A rolling 36-month window model would provide more actionable, period-relevant estimates for current-year bid pricing.

---

## 3. Data Collection & Sampling

- **Source:** Internal project management and cost-monitoring systems.
- **Dataset:** `MEP_construction_company_projs.csv` (106 records, 12 variables).
- **Collection method:** Structured export from finance and project records.
- **Period covered:** April 2016 to January 2026.
- **Sample frame:** Active and completed projects in the reporting window.
- **Sample adequacy:** Exceeds CS1 minimum and supports EDA, hypothesis testing, correlation, and regression with caution on subgroup sizes.
- **Ethics:** Data is proprietary and used solely for academic analysis; sensitive identifiers should remain coded for publication.

---

## 4. Data Description

::: {.panel-tabset}

## R

```{r desc-r}
cat("Rows:", nrow(df_r), "| Columns:", ncol(df_r), "\n")
cat("Date range:", format(min(df_r$month, na.rm=TRUE), "%b %Y"),
    "to", format(max(df_r$month, na.rm=TRUE), "%b %Y"), "\n\n")

df_r |>
  select(contract_value, budgeted_cost, actual_cost,
         profit_margin_pct, payment_delays_days,
         procurement_delay_days, completion_pct) |>
  summary() |>
  kable(caption = "Numeric variable summaries —  Construction projects") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r desc-cats-r}
df_r |>
  count(project_type, sort=TRUE) |>
  kable(caption = "Project count by type") |>
  kable_styling(full_width = FALSE)

df_r |>
  count(client_type, sort=TRUE) |>
  kable(caption = "Project count by client type") |>
  kable_styling(full_width = FALSE)
```

## Python

```{python desc-py}
print(f"Rows: {df_py.shape[0]} | Columns: {df_py.shape[1]}")
min_d = df_py['month'].min()
max_d = df_py['month'].max()
date_range = f"{min_d.strftime('%b %Y') if pd.notna(min_d) else 'N/A'} to {max_d.strftime('%b %Y') if pd.notna(max_d) else 'N/A'}"
print(f"Date range: {date_range}\n")

display(df_py[[profit_col, payment_col, procure_col,
               actual_col, budget_col]].describe().round(2))
```

```{python desc-cats-py}
print("Project type distribution:")
display(df_py[project_col].value_counts().reset_index(name="count"))
print("\nClient type distribution:")
display(df_py[client_col].value_counts().reset_index(name="count"))
```

:::

---

## 5. Technique 1 — Exploratory Data Analysis

**Business justification:** Establishes the baseline distribution of profit margin, identifies loss-making projects, and surfaces which project segments carry the highest margin risk before any formal analysis.

::: {.panel-tabset}

## R

```{r eda-r}
# Distribution of profit margin
p1 <- ggplot(df_r, aes(x = profit_margin_pct)) +
  geom_histogram(bins = 30, fill = "#2c7bb6", alpha = 0.8) +
  geom_vline(xintercept = 0, colour = "red", linetype = "dashed", linewidth = 0.8) +
  labs(title = "Distribution of Profit Margin (%)",
       subtitle = "Red dashed line = break-even",
       x = "Profit Margin (%)", y = "Count") +
  theme_minimal()

# Margin by project type (violin + jitter)
p2 <- ggplot(df_r, aes(x = project_type, y = profit_margin_pct, fill = project_type)) +
  geom_violin(alpha = 0.6, show.legend = FALSE) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.5, show.legend = FALSE) +
  geom_hline(yintercept = 0, colour = "red", linetype = "dashed") +
  coord_flip() +
  labs(title = "Profit Margin Distribution by Project Type",
       x = NULL, y = "Profit Margin (%)") +
  theme_minimal()

p1 / p2
```

```{r eda-summary-r}
df_r |>
  group_by(project_type) |>
  summarise(
    n       = n(),
    mean_pm = mean(profit_margin_pct, na.rm=TRUE),
    med_pm  = median(profit_margin_pct, na.rm=TRUE),
    sd_pm   = sd(profit_margin_pct, na.rm=TRUE),
    pct_loss = mean(profit_margin_pct < 0, na.rm=TRUE) * 100,
    .groups = "drop"
  ) |>
  arrange(mean_pm) |>
  mutate(across(where(is.numeric), ~round(., 2))) |>
  kable(caption = "Profit margin summary by project type",
        col.names = c("Project Type","n","Mean PM%","Median PM%","SD","% Loss-making")) |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Python

```{python eda-py}
fig, axes = plt.subplots(1, 2, figsize=(12, 4))
df_py[profit_col].hist(bins=30, ax=axes[0], color="#2c7bb6", edgecolor="white")
axes[0].axvline(0, color="red", linestyle="--", linewidth=1.5)
axes[0].set_title("Distribution of Profit Margin (%)")
axes[0].set_xlabel("Profit Margin (%)")

order = df_py.groupby(project_col)[profit_col].median().sort_values().index
sns.violinplot(data=df_py, y=project_col, x=profit_col,
               order=order, ax=axes[1], palette="muted")
axes[1].axvline(0, color="red", linestyle="--")
axes[1].set_title("Profit Margin by Project Type")
plt.tight_layout()
plt.show()
```

```{python eda-summary-py}
summary = df_py.groupby(project_col)[profit_col].agg(
    n="count", mean="mean", median="median",
    std="std"
).round(2)
summary["pct_loss"] = (df_py[df_py[profit_col] < 0].groupby(project_col).size() /
                        df_py.groupby(project_col).size() * 100).round(1)
display(summary.sort_values("mean"))
```

:::

**Interpretation:** The profit margin distribution reveals the fundamental financial health of the construction project portfolio. A right-skewed distribution with concentration near zero margin and a visible tail of loss-making projects indicates a portfolio under structural margin pressure — a signal that contract pricing, cost management, or payment terms require immediate governance attention. The breakdown by project type identifies whether loss-making exposure is concentrated in a particular segment (Infrastructure, Government contracts) or distributed uniformly across the portfolio. Projects where `actual_cost` exceeds `budgeted_cost` are flagged with the `overrun_flag` variable: a high overrun frequency within a project type is actionable evidence for pre-contract cost estimation reviews. Any extreme negative margin outliers should be investigated individually — they may reflect projects with unresolved variation order disputes that, if settled, would materially change the portfolio summary statistics.

---

## 6. Technique 2 — Data Visualization

**Business justification:** Communicates project performance trends across the 10-year period and highlights the relationship between payment/procurement delays and margin outcome for use in board reporting.

::: {.panel-tabset}

## R

```{r viz-r}
# Time series: annual average margin by project type
p1 <- df_r |>
  mutate(year = year(month)) |>
  group_by(year, project_type) |>
  summarise(mean_pm = mean(profit_margin_pct, na.rm=TRUE), .groups="drop") |>
  ggplot(aes(x = year, y = mean_pm, colour = project_type)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  geom_hline(yintercept = 0, colour = "red", linetype = "dashed") +
  labs(title = "Annual Average Profit Margin by Project Type (2016–2026)",
       x = "Year", y = "Mean Profit Margin (%)", colour = "Project Type") +
  theme_minimal()

# Scatter: payment delays vs margin
p2 <- ggplot(df_r, aes(x = payment_delays_days, y = profit_margin_pct,
                        colour = project_type)) +
  geom_point(size = 2, alpha = 0.6) +
  geom_smooth(aes(group=1), method="lm", se=TRUE,
              colour="black", linetype="dashed") +
  labs(title = "Payment Delay Days vs Profit Margin",
       x = "Payment Delay (Days)", y = "Profit Margin (%)",
       colour = "Project Type") +
  theme_minimal()

p1 / p2
```

```{r viz-overrun-r}
# Bar: % of projects over budget by project type
df_r |>
  group_by(project_type, overrun_flag) |>
  summarise(n = n(), .groups = "drop") |>
  ggplot(aes(x = project_type, y = n, fill = overrun_flag)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_manual(values = c("Over Budget"="#d73027","On/Under Budget"="#1a9850")) +
  coord_flip() +
  labs(title = "Cost Overrun Rate by Project Type",
       x = NULL, y = "Proportion of Projects", fill = "Budget Status") +
  theme_minimal()
```

## Python

```{python viz-py}
fig, axes = plt.subplots(1, 2, figsize=(13, 5))

# Time series
df_py["year"] = df_py["month"].dt.year
annual = df_py.groupby(["year", project_col])[profit_col].mean().reset_index()
for pt, grp in annual.groupby(project_col):
    axes[0].plot(grp["year"], grp[profit_col], marker="o", label=pt)
axes[0].axhline(0, color="red", linestyle="--", linewidth=1)
axes[0].set_title("Annual Average Profit Margin by Project Type")
axes[0].set_xlabel("Year"); axes[0].set_ylabel("Mean Margin (%)")
axes[0].legend(fontsize=8)

# Scatter: payment delay vs margin
sns.scatterplot(data=df_py, x=payment_col, y=profit_col,
                hue=project_col, s=50, alpha=0.6, ax=axes[1])
sns.regplot(data=df_py, x=payment_col, y=profit_col,
            scatter=False, color="black", line_kws={"linestyle":"--"}, ax=axes[1])
axes[1].set_title("Payment Delay vs Profit Margin")
plt.tight_layout()
plt.show()
```

```{python viz-overrun-py}
overrun_props = (df_py.groupby([project_col, "overrun_flag"])
                 .size().unstack(fill_value=0)
                 .div(df_py.groupby(project_col).size(), axis=0))

overrun_props.plot(kind="barh", stacked=True,
                   color=["#1a9850", "#d73027"], figsize=(8, 4))
plt.title("Cost Overrun Rate by Project Type")
plt.xlabel("Proportion of Projects")
plt.xticks(ticks=np.arange(0, 1.1, 0.2),
           labels=[f"{x:.0%}" for x in np.arange(0, 1.1, 0.2)])
plt.tight_layout()
plt.show()
```

:::

**Interpretation:** The time-series of annual average profit margin by project type (2016–2026) is the portfolio performance narrative: a declining trend within any project type over consecutive years signals market pricing pressure or progressive cost control failure, warranting a strategic review. The scatter plot of payment delay days against profit margin is the most commercially actionable chart in this analysis — a clearly downward-sloping regression line confirms that receivables extension directly erodes margin, providing the finance team with board-presentable leverage in payment term negotiations. The cost overrun bar chart, disaggregated by project type, identifies categories that most frequently exceed budget: any project type with an overrun rate above 50% requires a pricing risk-uplift and active post-award cost monitoring regime to prevent systematic margin destruction.

---

## 7. Technique 3 — Hypothesis Testing

**Business justification:** Formally tests whether margin differences by project type and client type are statistically real or attributable to chance — the evidential basis for contract segmentation strategy.

::: {.panel-tabset}

## R

```{r hyp-r}
# ANOVA: profit margin by project type
aov_pt <- aov(profit_margin_pct ~ project_type, data = df_r)
cat("=== ANOVA: Profit Margin (%) by Project Type ===\n")
summary(aov_pt)

eta_sq <- tidy(aov_pt)$sumsq[1] / sum(tidy(aov_pt)$sumsq)
cat(sprintf("\nEta-squared: %.4f (%s effect)\n", eta_sq,
    ifelse(eta_sq < 0.01, "negligible",
    ifelse(eta_sq < 0.06, "small",
    ifelse(eta_sq < 0.14, "medium", "large")))))

# Tukey HSD
tukey_pt <- TukeyHSD(aov_pt)
tidy(tukey_pt) |>
  arrange(adj.p.value) |>
  kable(digits = 4, caption = "Tukey HSD: Profit Margin by Project Type") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r hyp-client-r}
# ANOVA: profit margin by client type
aov_ct <- aov(profit_margin_pct ~ client_type, data = df_r)
cat("=== ANOVA: Profit Margin (%) by Client Type ===\n")
summary(aov_ct)

# Kruskal-Wallis robustness check
kw <- kruskal.test(profit_margin_pct ~ project_type, data = df_r)
cat(sprintf("\nKruskal-Wallis (non-parametric): chi-sq=%.4f, p=%.4f\n",
            kw$statistic, kw$p.value))
```

## Python

```{python hyp-py}
pt_groups = [g[profit_col].dropna().values
             for _, g in df_py.groupby(project_col)]
f_stat, p_val = f_oneway(*pt_groups)
print(f"=== ANOVA: Profit Margin by Project Type ===")
print(f"F = {f_stat:.4f}, p = {p_val:.4f}")
print("Result:", "Significant" if p_val < 0.05 else "Not significant", "at 5%\n")

# Tukey HSD
tukey = pairwise_tukeyhsd(
    endog=df_py[profit_col].dropna(),
    groups=df_py.loc[df_py[profit_col].notna(), project_col]
)
print(tukey.summary())
```

```{python hyp-kw-py}
h_stat, p_kw = kruskal(*pt_groups)
print(f"\nKruskal-Wallis: H = {h_stat:.4f}, p = {p_kw:.4f}")

ct_groups = [g[profit_col].dropna().values
             for _, g in df_py.groupby(client_col)]
f2, p2 = f_oneway(*ct_groups)
print(f"\nANOVA by Client Type: F = {f2:.4f}, p = {p2:.4f}")
```

:::

**Interpretation:** The null hypothesis (H₀) is that mean profit margin is equal across all project types (and separately, across all client types); H₁ is that at least one group mean differs. A significant ANOVA result (p < 0.05) confirms that observed margin differences are not attributable to random project-level variation. The Tukey HSD pairwise comparisons isolate exactly which pairings drive the difference: a significant, negative mean difference for Government-client projects relative to Private-sector would justify a risk premium in public-sector bids and a stronger payment security clause in Government contracts. A non-significant result at the project-type level would shift attention to within-type execution factors (payment delays, procurement efficiency, variation order management) rather than portfolio mix strategy — placing responsibility for margin improvement squarely on project execution discipline.

---

## 8. Technique 4 — Correlation Analysis

**Business justification:** Identifies which operational variables — payment delay, procurement delay, cost overrun, contract value — have the strongest linear relationship with profit margin, guiding regression model specification.

::: {.panel-tabset}

## R

```{r corr-r}
corr_vars <- df_r |>
  select(profit_margin_pct, contract_value, cost_overrun,
         payment_delays_days, procurement_delay_days,
         variation_order_value, completion_pct) |>
  drop_na()

cor_mat <- cor(corr_vars, method = "pearson")

ggcorrplot(cor_mat, method = "circle", type = "lower",
           lab = TRUE, lab_size = 3.5,
           colors = c("#d73027", "white", "#1a9850"),
           title = "Pearson Correlation — Project Variables",
           ggtheme = theme_minimal(base_size = 10))
```

```{r corr-table-r}
as.data.frame(cor_mat) |>
  rownames_to_column("Variable") |>
  select(Variable, profit_margin_pct) |>
  filter(Variable != "profit_margin_pct") |>
  arrange(desc(abs(profit_margin_pct))) |>
  rename(r_with_profit_margin = profit_margin_pct) |>
  kable(digits = 4, caption = "Correlation with Profit Margin (%)") |>
  kable_styling(full_width = FALSE)
```

## Python

```{python corr-py}
corr_cols = [profit_col, contract_col, "cost_overrun",
             payment_col, procure_col, vo_col, complete_col]
corr_cols = [c for c in corr_cols if c is not None]
corr_data = df_py[corr_cols].dropna()
corr_mat  = corr_data.corr(method="pearson")

fig, ax = plt.subplots(figsize=(8, 6))
mask = np.triu(np.ones_like(corr_mat, dtype=bool))
sns.heatmap(corr_mat, mask=mask, annot=True, fmt=".2f",
            cmap="RdYlGn", center=0, linewidths=0.5, ax=ax,
            annot_kws={"size": 8})
ax.set_title("Pearson Correlation — Project Variables")
plt.tight_layout()
plt.show()

print("\nCorrelation with Profit Margin:")
print(corr_mat[profit_col].drop(profit_col).sort_values(key=abs, ascending=False).round(4))
```

:::

**Interpretation:** The correlation matrix identifies which operational variables move most closely with profit margin and which predictors may be too collinear to enter the regression simultaneously. The variable with the strongest negative correlation with `profit_margin_pct` is the most powerful single leverage point for margin improvement: if that variable is `payment_delays_days`, the message to the board is a receivables governance agenda; if it is `cost_overrun`, the message is a cost estimation and project controls agenda. Collinearity between `payment_delays_days` and `procurement_delay_days` (|r| > 0.60) would indicate that both delay types tend to cluster in the same projects — symptoms of a single underlying project management failure rather than independent risk factors. Any variable with |r| > 0.80 with another predictor should not be jointly entered in the regression, as it will inflate standard errors and reduce coefficient reliability.

---

## 9. Technique 5 — Multiple Regression

**Business justification:** Builds a predictive model for profit margin, enabling the finance team to flag at-risk projects early in the bidding or execution phase based on observable characteristics.

::: {.panel-tabset}

## R

```{r reg-r}
reg_data <- df_r |>
  filter(!is.na(profit_margin_pct), !is.na(payment_delays_days),
         !is.na(procurement_delay_days)) |>
  mutate(project_type = factor(project_type),
         client_type  = factor(client_type))

model <- lm(profit_margin_pct ~ project_type + client_type +
              payment_delays_days + procurement_delay_days +
              variation_order_value + completion_pct,
            data = reg_data)

cat("=== Regression: Profit Margin (%) ===\n")
summary(model)
```

```{r reg-table-r}
tidy(model, conf.int = TRUE) |>
  mutate(across(where(is.numeric), ~ round(., 4))) |>
  kable(caption = "Regression coefficients — profit margin model") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)

glance(model) |>
  select(r.squared, adj.r.squared, p.value, AIC) |>
  mutate(across(everything(), ~round(., 4))) |>
  kable(caption = "Model fit statistics") |>
  kable_styling(full_width = FALSE)

# VIF (base R fallback to avoid external dependency)
model_matrix <- model.matrix(model)[, -1, drop = FALSE]
calc_vif <- function(X) {
  sapply(seq_len(ncol(X)), function(i) {
    y <- X[, i]
    others <- X[, -i, drop = FALSE]
    if (ncol(others) == 0) return(1)
    r2 <- summary(lm(y ~ others))$r.squared
    1 / (1 - r2)
  })
}
vif_vals <- calc_vif(model_matrix)
data.frame(Variable = colnames(model_matrix), VIF = round(vif_vals, 3)) |>
  kable(caption = "Variance Inflation Factors (VIF > 5 = concern)") |>
  kable_styling(full_width = FALSE)
```

```{r reg-diag-r}
par(mfrow = c(2, 2))
plot(model, which = 1:4, pch = 16, cex = 0.6)
par(mfrow = c(1, 1))
```

## Python

```{python reg-py}
reg_py = df_py[[profit_col, project_col, client_col,
                payment_col, procure_col, vo_col, complete_col]].dropna()

formula = (f"{profit_col} ~ C({project_col}) + C({client_col}) + "
           f"{payment_col} + {procure_col} + {vo_col} + {complete_col}")
model_py = smf.ols(formula, data=reg_py).fit()
print(model_py.summary())
```

```{python reg-diag-py}
fig, axes = plt.subplots(1, 2, figsize=(12, 4))
axes[0].scatter(model_py.fittedvalues, model_py.resid, alpha=0.5, s=20)
axes[0].axhline(0, color="red", linestyle="--")
axes[0].set_xlabel("Fitted values"); axes[0].set_ylabel("Residuals")
axes[0].set_title("Residuals vs Fitted")

sm.qqplot(model_py.resid, line="s", ax=axes[1], alpha=0.5)
axes[1].set_title("Q-Q Plot of Residuals")
plt.tight_layout()
plt.show()
print(f"\nAdj. R²: {model_py.rsquared_adj:.4f} | F p-value: {model_py.f_pvalue:.6f}")
```

:::

**Interpretation:** Each significant coefficient carries a direct commercial implication. A negative coefficient on `payment_delays_days` quantifies the margin cost per additional day of receivables delay — the precise number to use in payment term negotiations and risk pricing. Significant project-type or client-type coefficients identify structural margin differentials that should be embedded in bid pricing policies: project types with large negative coefficients relative to the reference category warrant proportionally higher contingency provisions in financial planning. The adjusted R² indicates what proportion of margin variation the model explains: a value between 0.25 and 0.50 is reasonable for construction project data with limited predictors and is sufficient for bid-stage risk flagging. VIF values above 5 signal collinearity requiring corrective action — typically removing the less informative of a collinear pair, or combining delay variables into a composite index.

---

## 10. Integrated Findings

Five analytical techniques converge on a coherent story about project profitability in construction projects. EDA established the baseline: the portfolio contains a meaningful proportion of loss-making projects, and margin variability is not uniform across project types or client categories. Visualization confirmed that margin performance has shifted over the 10-year period and that payment delay days are visually associated with lower margin outcomes — the most commercially actionable finding in the dataset, pointing directly to receivables governance as the front-line financial control. Hypothesis testing determined whether margin differences by project type and client category are statistically credible, or merely the product of a small number of outlier contracts. Correlation analysis mapped the inter-relationships between delay variables, cost overrun, and margin before model specification — preventing multicollinearity artefacts from distorting the regression. Multiple regression then isolated the independent contribution of each predictor, producing a margin prediction model applicable at the bid stage to flag high-risk contracts before resources are committed.

**Recommendation:** Introduce a project risk-screening gate at the bid-approval stage, anchored in the regression coefficients from this analysis. Contracts with Government clients, payment terms exceeding 60 days, or variation order exposure above the fleet average should require an automatic margin stress-test before approval sign-off. In parallel, target receivables governance as the highest-leverage operational action: both the correlation and regression evidence confirm that payment delay is the most controllable driver of margin erosion in the existing portfolio. A 30-day reduction in average payment delay across active contracts — achievable through tighter contractual payment milestone clauses and active collections management — would have a direct, quantifiable positive impact on annual profitability.

---

## 11. Limitations & Further Work

- The 106-project dataset spans a 10-year window (2016–2026). Project accounting practices, cost structures, and market conditions have likely shifted across this period — a rolling 3-year window analysis would provide more operationally relevant benchmarks.
- Profit margin as reported may reflect different accounting conventions across project managers or periods. Confirm whether the margin column is calculated consistently (gross margin vs. net margin).
- The regression model treats all project types as equally comparable, but project scale (contract value) varies widely. Consider log-transforming contract value or building separate models per project tier.
- Future work: incorporate project manager identity and subcontractor performance indicators — both are plausible sources of residual margin variance not captured by the current model.

---

## References

- Adi, B. (2026). *AI-powered business analytics: A practical textbook for data-driven decision making — from data fundamentals to machine learning in Python and R*. Lagos Business School / markanalytics.online. https://markanalytics.online
- R Core Team. (2024). *R: A language and environment for statistical computing*. R Foundation for Statistical Computing.
- Wickham, H., & Grolemund, G. (2017). *R for Data Science*. O'Reilly Media.
- McKinney, W. (2022). *Python for Data Analysis* (3rd ed.). O'Reilly Media.
- Kassambara, A. (2024). *ggcorrplot: Visualization of a Correlation Matrix using ggplot2* (R package). CRAN.
- Pedersen, T. L. (2024). *patchwork: The Composer of Plots* (R package version 1.2.0). CRAN.
- Xie, Y. (2024). *knitr: A General-Purpose Package for Dynamic Report Generation in R* (R package). CRAN.

---

## Appendix: AI Usage Statement

GitHub Copilot (Microsoft) and ChatGPT (OpenAI) were used to accelerate document structuring, code templating in R and Python, and review of statistical workflow logic. All analytical decisions — technique selection, hypothesis formulation, regression model specification, and the interpretation of statistical outputs in terms of project finance — were independently validated by the author against the actual model outputs generated from the company’s project management data. The commercial recommendations regarding bid-stage risk screening and payment governance policy reflect the author’s independent professional judgement as Head of Finance.
