---
title: "Project Profitability Analytics: Drivers of Profit Margin in MEP Construction Projects"
author: "Oladayo Folorunso"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
execute:
  warning: false
  message: false
---

```{r setup, include=FALSE}
pkgs <- c("tidyverse", "lubridate", "scales", "ggcorrplot",
          "patchwork", "car", "broom", "knitr", "kableExtra", "janitor")
invisible(lapply(pkgs, library, character.only = TRUE))
knitr::opts_chunk$set(fig.width = 9, fig.height = 5, dpi = 150)
```

```{python setup-py, include=FALSE}
import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
import matplotlib.ticker as mticker
import seaborn as sns
from scipy import stats
from scipy.stats import f_oneway, kruskal, ttest_ind, pearsonr
import statsmodels.formula.api as smf
import statsmodels.api as sm
from statsmodels.stats.multicomp import pairwise_tukeyhsd
from IPython.display import display
sns.set_theme(style="whitegrid", palette="muted")
plt.rcParams["figure.dpi"] = 150
```

```{r load-r, include=FALSE}
# ⚠ Column names in raw CSV have leading/trailing whitespace —
# janitor::clean_names() resolves all of these automatically.
df_r <- read_csv("data/MEP_construction_company_projs.csv",
                 col_types = cols(.default = col_character())) |>
  clean_names() |>    # strips whitespace, lowercases, converts spaces → underscores
  mutate(
    month               = my(month),           # "Apr-16" → Date
    contract_value      = as.numeric(gsub("[^0-9.]", "", contract_value)),
    budgeted_cost       = as.numeric(gsub("[^0-9.]", "", budgeted_cost)),
    actual_cost         = as.numeric(gsub("[^0-9.]", "", actual_cost)),
    profit_margin_pct   = as.numeric(gsub("[^0-9.\\-]", "", profit_margin_percent)),
    payment_delays_days = as.numeric(payment_delays_days),
    variation_order_value    = as.numeric(gsub("[^0-9.\\-]", "", variation_order_value)),
    procurement_delay_days   = as.numeric(procurement_delay_days),
    completion_pct           = as.numeric(gsub("[^0-9.]", "", completion_percent))
  ) |>
  filter(!is.na(profit_margin_pct))

# Derived: cost overrun flag
df_r <- df_r |>
  mutate(cost_overrun = actual_cost - budgeted_cost,
         overrun_flag = if_else(cost_overrun > 0, "Over Budget", "On/Under Budget"))
```

```{python load-py, include=FALSE}
df_py = pd.read_csv("data/MEP_construction_company_projs.csv", dtype=str)

# ⚠ Strip leading/trailing whitespace from all column names
df_py.columns = df_py.columns.str.strip()

# Normalise to snake_case
df_py.columns = (df_py.columns.str.lower()
                 .str.replace(r"[^\w]", "_", regex=True)
                 .str.replace(r"_+", "_", regex=True)
                 .str.strip("_"))

# Parse month
df_py["month"] = pd.to_datetime(df_py["month"].str.strip(), format="%b-%y", errors="coerce")

# Numeric columns
num_cols = ["contract_value", "budgeted_cost", "actual_cost",
            "profit_margin", "payment_delays_days",
            "variation_order_value", "procurement_delay_days", "completion"]
for col in num_cols:
    # match against clean column name
    match = [c for c in df_py.columns if col.replace("_", "") in c.replace("_", "")]
    if match:
        df_py[match[0]] = pd.to_numeric(
            df_py[match[0]].str.replace(r"[^\d.\-]", "", regex=True), errors="coerce")

# Re-identify key columns by partial match for safety
def find_col(df, keyword):
    matches = [c for c in df.columns if keyword in c]
    return matches[0] if matches else None

profit_col   = find_col(df_py, "profit_margin")
payment_col  = find_col(df_py, "payment_delay")
procure_col  = find_col(df_py, "procurement_delay")
actual_col   = find_col(df_py, "actual_cost")
budget_col   = find_col(df_py, "budgeted_cost")
contract_col = find_col(df_py, "contract_value")
project_col  = find_col(df_py, "project_type")
client_col   = find_col(df_py, "client_type")
complete_col = find_col(df_py, "completion")
vo_col       = find_col(df_py, "variation_order")

df_py = df_py.dropna(subset=[profit_col])
df_py["cost_overrun"] = df_py[actual_col] - df_py[budget_col]
df_py["overrun_flag"] = np.where(df_py["cost_overrun"] > 0, "Over Budget", "On/Under Budget")
```

---

## 1. Executive Summary

> **Instructions:** Complete after your analysis. 150–200 words.

**Business problem:** Profit margin variability is the central financial risk for MEP construction project delivery. Understanding which project characteristics — type, client category, payment behaviour, procurement delays — drive margin erosion enables the finance function to price contracts more accurately and implement early-warning controls during execution.

**Data collected:** 106 MEP construction projects executed between April 2016 and January 2026, drawn from the company's project management information system. Metrics include contract value, budgeted and actual cost, profit margin (%), payment delays, variation orders, procurement delays, and project completion percentage.

**Key findings:**

1. [Project Type X] projects deliver a median margin of [Y%], compared to [Z%] for [Type W] — a statistically significant difference (p = [value]).
2. Payment delay days show a [negative] correlation with profit margin (r = [value]).
3. The regression model explains [R²]% of margin variance, identifying [top driver] as the primary lever.

**Recommendation:** [Specific policy action with projected margin impact.]

---

## 2. Professional Disclosure

> **Instructions:** Paste your completed Section 2 from `docs/section_a_30_marks_template.md` here.

---

## 3. Data Collection & Sampling

> **Instructions:** Paste your completed Section 3 from `docs/section_a_30_marks_template.md` here.

---

## 4. Data Description

::: {.panel-tabset}

## R

```{r desc-r}
cat("Rows:", nrow(df_r), "| Columns:", ncol(df_r), "\n")
cat("Date range:", format(min(df_r$month, na.rm=TRUE), "%b %Y"),
    "to", format(max(df_r$month, na.rm=TRUE), "%b %Y"), "\n\n")

df_r |>
  select(contract_value, budgeted_cost, actual_cost,
         profit_margin_pct, payment_delays_days,
         procurement_delay_days, completion_pct) |>
  summary() |>
  kable(caption = "Numeric variable summaries — MEP construction projects") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r desc-cats-r}
df_r |>
  count(project_type, sort=TRUE) |>
  kable(caption = "Project count by type") |>
  kable_styling(full_width = FALSE)

df_r |>
  count(client_type, sort=TRUE) |>
  kable(caption = "Project count by client type") |>
  kable_styling(full_width = FALSE)
```

## Python

```{python desc-py}
print(f"Rows: {df_py.shape[0]} | Columns: {df_py.shape[1]}")
print(f"Date range: {df_py['month'].min().strftime('%b %Y')} "
      f"to {df_py['month'].max().strftime('%b %Y')}\n")

display(df_py[[profit_col, payment_col, procure_col,
               actual_col, budget_col]].describe().round(2))
```

```{python desc-cats-py}
print("Project type distribution:")
display(df_py[project_col].value_counts().reset_index(name="count"))
print("\nClient type distribution:")
display(df_py[client_col].value_counts().reset_index(name="count"))
```

:::

---

## 5. Technique 1 — Exploratory Data Analysis

**Business justification:** Establishes the baseline distribution of profit margin, identifies loss-making projects, and surfaces which project segments carry the highest margin risk before any formal analysis.

::: {.panel-tabset}

## R

```{r eda-r}
# Distribution of profit margin
p1 <- ggplot(df_r, aes(x = profit_margin_pct)) +
  geom_histogram(bins = 30, fill = "#2c7bb6", alpha = 0.8) +
  geom_vline(xintercept = 0, colour = "red", linetype = "dashed", linewidth = 0.8) +
  labs(title = "Distribution of Profit Margin (%)",
       subtitle = "Red dashed line = break-even",
       x = "Profit Margin (%)", y = "Count") +
  theme_minimal()

# Margin by project type (violin + jitter)
p2 <- ggplot(df_r, aes(x = project_type, y = profit_margin_pct, fill = project_type)) +
  geom_violin(alpha = 0.6, show.legend = FALSE) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.5, show.legend = FALSE) +
  geom_hline(yintercept = 0, colour = "red", linetype = "dashed") +
  coord_flip() +
  labs(title = "Profit Margin Distribution by Project Type",
       x = NULL, y = "Profit Margin (%)") +
  theme_minimal()

p1 / p2
```

```{r eda-summary-r}
df_r |>
  group_by(project_type) |>
  summarise(
    n       = n(),
    mean_pm = mean(profit_margin_pct, na.rm=TRUE),
    med_pm  = median(profit_margin_pct, na.rm=TRUE),
    sd_pm   = sd(profit_margin_pct, na.rm=TRUE),
    pct_loss = mean(profit_margin_pct < 0, na.rm=TRUE) * 100,
    .groups = "drop"
  ) |>
  arrange(mean_pm) |>
  mutate(across(where(is.numeric), ~round(., 2))) |>
  kable(caption = "Profit margin summary by project type",
        col.names = c("Project Type","n","Mean PM%","Median PM%","SD","% Loss-making")) |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Python

```{python eda-py}
fig, axes = plt.subplots(1, 2, figsize=(12, 4))
df_py[profit_col].hist(bins=30, ax=axes[0], color="#2c7bb6", edgecolor="white")
axes[0].axvline(0, color="red", linestyle="--", linewidth=1.5)
axes[0].set_title("Distribution of Profit Margin (%)")
axes[0].set_xlabel("Profit Margin (%)")

order = df_py.groupby(project_col)[profit_col].median().sort_values().index
sns.violinplot(data=df_py, y=project_col, x=profit_col,
               order=order, ax=axes[1], palette="muted")
axes[1].axvline(0, color="red", linestyle="--")
axes[1].set_title("Profit Margin by Project Type")
plt.tight_layout()
plt.show()
```

```{python eda-summary-py}
summary = df_py.groupby(project_col)[profit_col].agg(
    n="count", mean="mean", median="median",
    std="std"
).round(2)
summary["pct_loss"] = (df_py[df_py[profit_col] < 0].groupby(project_col).size() /
                        df_py.groupby(project_col).size() * 100).round(1)
display(summary.sort_values("mean"))
```

:::

> **Interpretation:** What is the overall margin profile — what % of projects are loss-making? Which project type has the most variable margin? Identify any outlier projects with unusually large negative margins.

---

## 6. Technique 2 — Data Visualization

**Business justification:** Communicates project performance trends across the 10-year period and highlights the relationship between payment/procurement delays and margin outcome for use in board reporting.

::: {.panel-tabset}

## R

```{r viz-r}
# Time series: annual average margin by project type
p1 <- df_r |>
  mutate(year = year(month)) |>
  group_by(year, project_type) |>
  summarise(mean_pm = mean(profit_margin_pct, na.rm=TRUE), .groups="drop") |>
  ggplot(aes(x = year, y = mean_pm, colour = project_type)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  geom_hline(yintercept = 0, colour = "red", linetype = "dashed") +
  labs(title = "Annual Average Profit Margin by Project Type (2016–2026)",
       x = "Year", y = "Mean Profit Margin (%)", colour = "Project Type") +
  theme_minimal()

# Scatter: payment delays vs margin
p2 <- ggplot(df_r, aes(x = payment_delays_days, y = profit_margin_pct,
                        colour = project_type)) +
  geom_point(size = 2, alpha = 0.6) +
  geom_smooth(aes(group=1), method="lm", se=TRUE,
              colour="black", linetype="dashed") +
  labs(title = "Payment Delay Days vs Profit Margin",
       x = "Payment Delay (Days)", y = "Profit Margin (%)",
       colour = "Project Type") +
  theme_minimal()

p1 / p2
```

```{r viz-overrun-r}
# Bar: % of projects over budget by project type
df_r |>
  group_by(project_type, overrun_flag) |>
  summarise(n = n(), .groups = "drop") |>
  ggplot(aes(x = project_type, y = n, fill = overrun_flag)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_manual(values = c("Over Budget"="#d73027","On/Under Budget"="#1a9850")) +
  coord_flip() +
  labs(title = "Cost Overrun Rate by Project Type",
       x = NULL, y = "Proportion of Projects", fill = "Budget Status") +
  theme_minimal()
```

## Python

```{python viz-py}
fig, axes = plt.subplots(1, 2, figsize=(13, 5))

# Time series
df_py["year"] = df_py["month"].dt.year
annual = df_py.groupby(["year", project_col])[profit_col].mean().reset_index()
for pt, grp in annual.groupby(project_col):
    axes[0].plot(grp["year"], grp[profit_col], marker="o", label=pt)
axes[0].axhline(0, color="red", linestyle="--", linewidth=1)
axes[0].set_title("Annual Average Profit Margin by Project Type")
axes[0].set_xlabel("Year"); axes[0].set_ylabel("Mean Margin (%)")
axes[0].legend(fontsize=8)

# Scatter: payment delay vs margin
sns.scatterplot(data=df_py, x=payment_col, y=profit_col,
                hue=project_col, s=50, alpha=0.6, ax=axes[1])
sns.regplot(data=df_py, x=payment_col, y=profit_col,
            scatter=False, color="black", line_kws={"linestyle":"--"}, ax=axes[1])
axes[1].set_title("Payment Delay vs Profit Margin")
plt.tight_layout()
plt.show()
```

```{python viz-overrun-py}
overrun_props = (df_py.groupby([project_col, "overrun_flag"])
                 .size().unstack(fill_value=0)
                 .div(df_py.groupby(project_col).size(), axis=0))

overrun_props.plot(kind="barh", stacked=True,
                   color=["#1a9850", "#d73027"], figsize=(8, 4))
plt.title("Cost Overrun Rate by Project Type")
plt.xlabel("Proportion of Projects")
plt.xticks(ticks=np.arange(0, 1.1, 0.2),
           labels=[f"{x:.0%}" for x in np.arange(0, 1.1, 0.2)])
plt.tight_layout()
plt.show()
```

:::

> **Interpretation:** Is there a visible trend in margin over the 10-year period? Does payment delay have an obvious negative relationship with margin? Which project type is most frequently over budget?

---

## 7. Technique 3 — Hypothesis Testing

**Business justification:** Formally tests whether margin differences by project type and client type are statistically real or attributable to chance — the evidential basis for contract segmentation strategy.

::: {.panel-tabset}

## R

```{r hyp-r}
# ANOVA: profit margin by project type
aov_pt <- aov(profit_margin_pct ~ project_type, data = df_r)
cat("=== ANOVA: Profit Margin (%) by Project Type ===\n")
summary(aov_pt)

eta_sq <- tidy(aov_pt)$sumsq[1] / sum(tidy(aov_pt)$sumsq)
cat(sprintf("\nEta-squared: %.4f (%s effect)\n", eta_sq,
    ifelse(eta_sq < 0.01, "negligible",
    ifelse(eta_sq < 0.06, "small",
    ifelse(eta_sq < 0.14, "medium", "large")))))

# Tukey HSD
tukey_pt <- TukeyHSD(aov_pt)
tidy(tukey_pt) |>
  arrange(adj.p.value) |>
  kable(digits = 4, caption = "Tukey HSD: Profit Margin by Project Type") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r hyp-client-r}
# ANOVA: profit margin by client type
aov_ct <- aov(profit_margin_pct ~ client_type, data = df_r)
cat("=== ANOVA: Profit Margin (%) by Client Type ===\n")
summary(aov_ct)

# Kruskal-Wallis robustness check
kw <- kruskal.test(profit_margin_pct ~ project_type, data = df_r)
cat(sprintf("\nKruskal-Wallis (non-parametric): chi-sq=%.4f, p=%.4f\n",
            kw$statistic, kw$p.value))
```

## Python

```{python hyp-py}
pt_groups = [g[profit_col].dropna().values
             for _, g in df_py.groupby(project_col)]
f_stat, p_val = f_oneway(*pt_groups)
print(f"=== ANOVA: Profit Margin by Project Type ===")
print(f"F = {f_stat:.4f}, p = {p_val:.4f}")
print("Result:", "Significant" if p_val < 0.05 else "Not significant", "at 5%\n")

# Tukey HSD
tukey = pairwise_tukeyhsd(
    endog=df_py[profit_col].dropna(),
    groups=df_py.loc[df_py[profit_col].notna(), project_col]
)
print(tukey.summary())
```

```{python hyp-kw-py}
h_stat, p_kw = kruskal(*pt_groups)
print(f"\nKruskal-Wallis: H = {h_stat:.4f}, p = {p_kw:.4f}")

ct_groups = [g[profit_col].dropna().values
             for _, g in df_py.groupby(client_col)]
f2, p2 = f_oneway(*ct_groups)
print(f"\nANOVA by Client Type: F = {f2:.4f}, p = {p2:.4f}")
```

:::

> **Interpretation:** State H₀ and H₁. For each significant test, write one business sentence. Example: "Government-client projects deliver a significantly lower median margin than private-sector projects (p = 0.03), suggesting a pricing or risk-uplift adjustment is warranted for public-sector bids."

---

## 8. Technique 4 — Correlation Analysis

**Business justification:** Identifies which operational variables — payment delay, procurement delay, cost overrun, contract value — have the strongest linear relationship with profit margin, guiding regression model specification.

::: {.panel-tabset}

## R

```{r corr-r}
corr_vars <- df_r |>
  select(profit_margin_pct, contract_value, cost_overrun,
         payment_delays_days, procurement_delay_days,
         variation_order_value, completion_pct) |>
  drop_na()

cor_mat <- cor(corr_vars, method = "pearson")

ggcorrplot(cor_mat, method = "circle", type = "lower",
           lab = TRUE, lab_size = 3.5,
           colors = c("#d73027", "white", "#1a9850"),
           title = "Pearson Correlation — MEP Project Variables",
           ggtheme = theme_minimal(base_size = 10))
```

```{r corr-table-r}
as.data.frame(cor_mat) |>
  rownames_to_column("Variable") |>
  select(Variable, profit_margin_pct) |>
  filter(Variable != "profit_margin_pct") |>
  arrange(desc(abs(profit_margin_pct))) |>
  rename(r_with_profit_margin = profit_margin_pct) |>
  kable(digits = 4, caption = "Correlation with Profit Margin (%)") |>
  kable_styling(full_width = FALSE)
```

## Python

```{python corr-py}
corr_cols = [profit_col, contract_col, "cost_overrun",
             payment_col, procure_col, vo_col, complete_col]
corr_cols = [c for c in corr_cols if c is not None]
corr_data = df_py[corr_cols].dropna()
corr_mat  = corr_data.corr(method="pearson")

fig, ax = plt.subplots(figsize=(8, 6))
mask = np.triu(np.ones_like(corr_mat, dtype=bool))
sns.heatmap(corr_mat, mask=mask, annot=True, fmt=".2f",
            cmap="RdYlGn", center=0, linewidths=0.5, ax=ax,
            annot_kws={"size": 8})
ax.set_title("Pearson Correlation — MEP Project Variables")
plt.tight_layout()
plt.show()

print("\nCorrelation with Profit Margin:")
print(corr_mat[profit_col].drop(profit_col).sort_values(key=abs, ascending=False).round(4))
```

:::

> **Interpretation:** Which variable correlates most strongly (negatively) with profit margin? Is there collinearity between payment delay and procurement delay? These findings should directly inform regression variable selection.

---

## 9. Technique 5 — Multiple Regression

**Business justification:** Builds a predictive model for profit margin, enabling the finance team to flag at-risk projects early in the bidding or execution phase based on observable characteristics.

::: {.panel-tabset}

## R

```{r reg-r}
reg_data <- df_r |>
  filter(!is.na(profit_margin_pct), !is.na(payment_delays_days),
         !is.na(procurement_delay_days)) |>
  mutate(project_type = factor(project_type),
         client_type  = factor(client_type))

model <- lm(profit_margin_pct ~ project_type + client_type +
              payment_delays_days + procurement_delay_days +
              variation_order_value + completion_pct,
            data = reg_data)

cat("=== Regression: Profit Margin (%) ===\n")
summary(model)
```

```{r reg-table-r}
tidy(model, conf.int = TRUE) |>
  mutate(across(where(is.numeric), ~ round(., 4))) |>
  kable(caption = "Regression coefficients — profit margin model") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)

glance(model) |>
  select(r.squared, adj.r.squared, p.value, AIC) |>
  mutate(across(everything(), ~round(., 4))) |>
  kable(caption = "Model fit statistics") |>
  kable_styling(full_width = FALSE)

# VIF
vif_vals <- car::vif(model)
data.frame(Variable = names(vif_vals), VIF = round(vif_vals, 3)) |>
  kable(caption = "Variance Inflation Factors (VIF > 5 = concern)") |>
  kable_styling(full_width = FALSE)
```

```{r reg-diag-r}
par(mfrow = c(2, 2))
plot(model, which = 1:4, pch = 16, cex = 0.6)
par(mfrow = c(1, 1))
```

## Python

```{python reg-py}
reg_py = df_py[[profit_col, project_col, client_col,
                payment_col, procure_col, vo_col, complete_col]].dropna()

formula = (f"{profit_col} ~ C({project_col}) + C({client_col}) + "
           f"{payment_col} + {procure_col} + {vo_col} + {complete_col}")
model_py = smf.ols(formula, data=reg_py).fit()
print(model_py.summary())
```

```{python reg-diag-py}
fig, axes = plt.subplots(1, 2, figsize=(12, 4))
axes[0].scatter(model_py.fittedvalues, model_py.resid, alpha=0.5, s=20)
axes[0].axhline(0, color="red", linestyle="--")
axes[0].set_xlabel("Fitted values"); axes[0].set_ylabel("Residuals")
axes[0].set_title("Residuals vs Fitted")

sm.qqplot(model_py.resid, line="s", ax=axes[1], alpha=0.5)
axes[1].set_title("Q-Q Plot of Residuals")
plt.tight_layout()
plt.show()
print(f"\nAdj. R²: {model_py.rsquared_adj:.4f} | F p-value: {model_py.f_pvalue:.6f}")
```

:::

> **Interpretation:** For each significant coefficient, write one business sentence. Example: "Each additional day of payment delay reduces expected project margin by [X] percentage points (β = −X, p < 0.05), implying that a payment terms policy capping delays at [N] days would protect approximately [Y%] of margin."

---

## 10. Integrated Findings

> Write 150–200 words connecting the five analyses into one coherent narrative and a single prioritized recommendation.

---

## 11. Limitations & Further Work

- The 106-project dataset spans a 10-year window (2016–2026). Project accounting practices, cost structures, and market conditions have likely shifted across this period — a rolling 3-year window analysis would provide more operationally relevant benchmarks.
- Profit margin as reported may reflect different accounting conventions across project managers or periods. Confirm whether the margin column is calculated consistently (gross margin vs. net margin).
- The regression model treats all project types as equally comparable, but project scale (contract value) varies widely. Consider log-transforming contract value or building separate models per project tier.
- Future work: incorporate project manager identity and subcontractor performance indicators — both are plausible sources of residual margin variance not captured by the current model.

---

## References

> *(Complete in APA format)*

---

## Appendix: AI Usage Statement

> *(One paragraph describing AI assistance and independent judgement.)*
